---
title: "The basics of transcriptomics pt3"
subtitle: "Differential Gene Experession Analysis and Visualization"
author: "Dr. Ido Hatam"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-copy: true
    self-contained: true
---

```{=html}
<style type="text/css">
body{ font-size: 18px; font-family: Arial, Helvetica, sans-serif; line-height: 30px; }
td { font-size: 14px; }
ul { font-size: 18px; }
h1.title { font-size: 38px; color: Black; }
h1 { font-size: 28px; color: Black; }
h2 { font-size: 22px; color: Black; }
h3 { font-size: 18px; font-family: Arial, Helvetica, sans-serif; color: Black; }
code.r{ font-size: 14px; }
.tocify { border: none; }
pre { font-size: 14px; }
</style>
```
# **The Final Frontier: Day 4 Overview**

You have spent the week building a high-performance bioinformatics pipeline. You have moved from raw, cryptic sequencing reads to a cleaned, normalized, and statistically rigorous list of genes that are significantly differentially expressed.

However, a list of Gene IDs like `AT3G12340` is not a discovery—it is a puzzle. Today, we solve that puzzle. We will move beyond the math to ask: **What is actually happening inside the plant?**

**Our goals for today:**

1.  **Global Visualization:** Use Volcano Plots to see the "magnitude vs. significance" landscape.

2.  **Pattern Recognition:** Use Heatmaps and Z-scores to visualize how gene expression fluctuates across time and treatments.

3.  **Functional Annotation:** Translate Gene IDs into biological pathways using Gene Ontology (GO) Enrichment.

4.  **Synthesis:** Interpret the temporal shifts (1h vs 12h and 1h vs 6h) to see how the plant's transcriptome changes over a half-day and see if we can still spot any infection markers.

# **Part 0: Computational Environment**

Today we require "Annotation Databases." These are massive dictionaries that map TAIR Gene IDs to their common names and biological functions.

1.  Create `transcriptomics_day4.qmd`.

2.  Load the following packages:

```{r}
#| message: false
#| warning: false

# Check and install missing critical packages if they don't exist
if (!requireNamespace("org.At.tair.db", quietly = TRUE)) BiocManager::install("org.At.tair.db", update = FALSE)
if (!requireNamespace("viridis", quietly = TRUE)) install.packages("viridis")

# Define the required packages
# clusterProfiler: For GO enrichment
# org.At.tair.db: The Arabidopsis annotation database
# viridis: For colorblind-friendly palettes
pkgs <- c("tidyverse", "DESeq2", "pheatmap", "EnhancedVolcano", 
          "clusterProfiler", "org.At.tair.db", "RColorBrewer", "viridis")

# Load all packages efficiently using lapply
lapply(pkgs, library, character.only = TRUE)

# Load your dds_results and count matrix from Day 3
# Ensure these objects are in your environment before proceeding
print("Biological Discovery Environment: READY")
```

# **Part 1: The Volcano Plot**

The Volcano Plot is a great way for visualization of DGE data. It allows us to see the entire transcriptome in a single image, highlighting the most extreme biological responders.

### **1. Theory: The Trade-off**

A Volcano Plot plots two variables:

-   **X-axis (**$log_2$ **Fold Change):** This represents the **Magnitude** of change.

-   **Y-axis (**$-log_{10}$ **p-value):** This represents the **Significance**.

**Why** $-log_{10}$**?** A tiny p-value (e.g., $0.00001$) is very significant. On a standard scale, these tiny numbers would all cluster at zero. By taking the negative log, we turn $10^{-5}$ into $5$. The higher the point on the Y-axis, the more confident we are that the change is real.

### **2. Example:1 2h vs. 1h Temporal Shif**

```{r}         
#| eval: false
#| label: vplot

EnhancedVolcano(res_1h_12h,
                lab = rownames(res_1h_12h),
                x = 'log2FoldChange',
                y = 'padj',
                title = 'Time Series: 12h vs. 1h',
                pCutoff = 0.2, 
                FCcutoff = 1.0,
                pointSize = 3.0,
                labSize = 4.0)
```

**Now You:** 

1. Generate a Volcano plot for your 1h vs. 6h results.

2. Visually identify if the "mountain" of genes looks broader or narrower than the 12h comparison.

**Bonus:** Do it for 6h vs 12h


# **Part 2: The Heatmap and Z-Scores**

A heatmap allows us to visualize how specific genes behave across **all** replicates and conditions simultaneously.

### **1. Theory: Why Z-Scores?**

If Gene A has 10,000 counts and Gene B has 10 counts, plotting them on the same map is impossible—Gene A will be bright red and Gene B will look black/empty.

To fix this, we use **Z-score normalization (Row Scaling)**. For every gene, we calculate:

$Z = \frac{x - mu}{sigma}$

A Z-score of **0** means the sample is exactly average. A Z-score of **+2** means the sample is much higher than its own average. This allows us to see **patterns of change** rather than raw magnitudes.

### **2. Theory: Understanding Dendrograms & Clustering**

When you generate a heatmap, you will see "trees" (lines) called **Dendrograms**.

-   **Hierarchical Clustering:** This is the algorithm that builds the trees. It groups the most similar items together.

-   **Row Clustering (Left):** This groups genes that have similar "behavior."

-   **Column Clustering (Top):** This groups your samples. This is a critical QC step—your biological replicates **must** cluster together.

### **3. Task: Visualizing the Top Responders (with Accessible Design)**

In data science, **accessibility matters**. Using "Red-Green" palettes is problematic for colorblindness. Even "Red-Blue" palettes can vary in readability. We will use the `viridis` palette, which is **colorblind-friendly** and remains interpretable even when printed in black and white.

```{r}         
#| eval: false
#| label: hmap

# 1. Extract the VST-transformed counts for your significant 1h vs 12h genes
# Assuming your filtered object is named 'sig_1h_12h'
plot_mat_12h <- assay(vsd)[rownames(sig_1h_12h), ]

# 2. Create the Heatmap
pheatmap(plot_mat_12h, 
         scale = "row", 
         clustering_distance_rows = "euclidean",
         clustering_method = "complete",
         color = magma(100), 
         show_colnames = TRUE,
         main = "Significant Genes: 12h vs 1h")
```

**Now You:** Create a second heatmap for the 1h vs 6h significant genes. Do you see the same genes in this comparison? Are the patterns as distinct?

**Bonus: The "Human Readable" Heatmap**

Working with AT3G12340 is technically correct but biologically uninspiring. We can use the org.At.tair.db to translate these IDs into Symbols (like CCA1).

Task: Re-run one of your heatmaps after running the following "translation" code:

```{r}
#| eval: false
#| label: hmap2

# 1. Map TAIR IDs to Symbols using the annotation database
gene_symbols <- mapIds(org.At.tair.db, 
                       keys = rownames(plot_mat_12h), 
                       column = "SYMBOL", 
                       keytype = "TAIR", 
                       multiVals = "first")

# 2. Replace the row names of your matrix
# We use ifelse to keep the TAIR ID if a Symbol doesn't exist
rownames(plot_mat_12h) <- ifelse(!is.na(gene_symbols), gene_symbols, rownames(plot_mat_12h))

# 3. Now re-run your pheatmap() call with plot_mat_12h!
```

**Search Challenge**: Look through your named heatmap for:

**Circadian Clock**: CCA1, TOC1, LHY.

**Photosynthesis**: PSB genes, CAB genes.

**Defense/PAMPs**: PR1, FLS2, WRKY transcription factors

# **Part 3: Gene Ontology (GO) Enrichment**

### **1. The Logic of Enrichment**

Imagine a bag of 30,000 marbles (the genome). 10% are blue (Defense genes). If you grab 100 marbles (your significant genes) and **80%** of them are blue, that category is **Enriched**.

### **2. Running enrichGO**

```{r}         
#| eval: false
#| label: enrich

# Select your significant UP-regulated genes from Avirulent
sig_avr_up <- res_avr_mock %>%
  as.data.frame() %>%
  filter(padj < 0.2 & log2FoldChange > 1) %>%
  rownames()

# Run the enrichment using TAIR IDs
ego_avr <- enrichGO(gene          = sig_avr_up,
                    OrgDb         = org.At.tair.db,
                    keyType       = "TAIR",
                    ont           = "BP", # Biological Process
                    pAdjustMethod = "BH",
                    pvalueCutoff  = 0.05,
                    qvalueCutoff  = 0.1)

# Visualize with a dotplot
# dotplot naturally uses the colorblind-friendly scales from ggplot2
enrichplot::dotplot(ego_avr, showCategory=20) + ggtitle("GO Enrichment: Avirulent Response")
```


**2. Pro Tip:** Custom Plotting with geom_point

If you want more control (e.g., changing colors or adding labels), you can convert the enrichResult object into a dataframe and use standard ggplot2.

```{r}
#| eval: false
#| label: dplot_geom

# 1. Convert to a dataframe
ego_df <- as.data.frame(ego_12h) %>%
  head(20) %>%
  # Secret trick: GeneRatio is a string "5/40". 
  # We use parse/eval to turn it into a decimal number (0.125).
  mutate(GeneRatioNumeric = sapply(GeneRatio, function(x) eval(parse(text=x))))

# 2. Plot with geom_point
ggplot(ego_df, aes(x = GeneRatioNumeric, y = reorder(Description, GeneRatioNumeric))) +
  geom_point(aes(size = Count, color = p.adjust)) +
  scale_color_viridis_c(option = "magma", direction = -1) +
  theme_minimal() +
  labs(x = "Gene Ratio", y = "Biological Process", 
       title = "Custom GO Enrichment Plot",
       size = "Gene Count", color = "Adj. P-value")
```


**Now You:** 
1. Perform GO enrichment for the 1h vs 6h comparison.
2. Use the geom_point method to create a custom plot for your 1h vs 6h results.
3. Compare the two plots. 
3. What are the top 3 biological processes the plant activates during the time since infection intervals?
4. Are there any pathways present in the **6h** response that are missing in the **12h** response?

# **Part 4: The Final Synthesis**

### **Final Project Task:**

In your `.qmd`, summarize your results.

-   How many genes did the plant activate in response to the two different pathogens?

-   Based on your GO enrichment, did the plant recognize the virulent pathogen as effectively as the avirulent one?

-   Reference one specific GO term or one specific gene from your Volcano plot to support your conclusion.

# **Day 4 Wrap-Up: From Data to Knowledge**

**Final Project Task:**

In your .qmd, summarize your results.

  *. How did the plant's transcriptome change between 1h vs. 6h and 1h vs. 12h? Is there a significant overlap in the genes identified across these shifts?

  *. Despite the strong temporal signal, can you still spot evidence of a Pathogen Response (infection markers like PR1 or WRKY) hidden in these lists?

  *. Reference one specific gene or GO term to support your conclusion.
  
# **Final Submission**


### **Git Final Push:**

Commit and push your final `.qmd`, `.html``.
