---
title: "Week 1, Day 2: The Notebook & The Backpack"
subtitle: "Tutorial & Assignment: Quarto, Lists, and Data Wrangling"
author: "Dr. Ido Hatam"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format: 
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-copy: true
    self-contained: true
---

```{=html}
<style type="text/css">

body{ /* Normal  */
      font-size: 18px;
      font-family: Arial, Helvetica, sans-serif;
      line-hight: 30px;
  }
td {  /* Table  */
  font-size: 14px;
}
ul { /* List */
  font-size: 18px;
}
h1.title {
  font-size: 38px;
  color: Black;
}
h1 { /* Header 1 */
  font-size: 28px;
  color: Black;
}
h2 { /* Header 2 */
    font-size: 22px;
  color: Black;
}
h3 { /* Header 3 */
  font-size: 18px;
  font-family: Arial, Helvetica, sans-serif;
  color: Black;
}

code.r{ /* Code block */
    font-size: 14px;
}
.tocify {
    border: none;
}

pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>
```
```{r setup}
#| include: false
library("tidyverse")
```

# **Part 1: The Lab Notebook (Quarto)**

## **What is Quarto?**

In the "old days" of bioinformatics, you might run code in a terminal, save the plot as an image file, and then copy that plot into a Word document to write your report. This is **error-prone** and **hard to reproduce**.

**Quarto (`.qmd`)** solves this by being a single file that contains:

1.  **Your Narrative:** The text explaining your science.

2.  **Your Code:** The R (or Python/Bash) commands that do the work.

3.  **Your Results:** The plots and tables generated by that code.

When you "Render" a Quarto file, it runs the code fresh every time. This guarantees that the figure in your report exactly matches the data you analyzed. It is the gold standard for **Reproducible Science**.

## **Starting a Quarto Document**

To start a new Quarto document in RStudio, follow these steps:

1.  Click on `File` -\> `New File` -\> `Quarto Document...`.

2.  A menu will pop up allowing you to add a Title and Author.

3.  Ensure "HTML" is selected as the output format (this is the default).

4.  Click "Create".

This will open a new file with a pre-filled template (the YAML header and some example text). You can delete the example text once we go over the structure of the document, but **keep the YAML header** at the top.

## **The Visual Editor**

Up until now, you may have used the Source Editor (where you see all the raw code).

RStudio has a Visual Editor button (Look for the "A" symbol or "Visual" button at the top left of this pane).

1.  **Switch to Visual Mode now.** \* Notice how the text looks like Microsoft Word?

    -   You can bold, italicize, and add headers without knowing Markdown syntax.

2.  **The Code Chunks:** \* Even in Visual Mode, the gray boxes remain code.

    -   You can run them using the green "Play" button.

## **The YAML Header**

Scroll to the very top of this file. You see data between three dashes ---.

This is the YAML. It controls how your document looks.

-   `toc: true` adds the Table of Contents.

-   `code-fold: show` allows readers to hide/show code.

## **Chunk Options**

Code chunks can be inserted using the code chunk button, or by using ```.
Once a code chunk has been initialized need to set the language for it to be evaluated properly
e.g. ```{r} will be evaluated as an R code. 

Inside a code chunk, we can control behavior using the hash-pipe `#|`.

-   `echo: true` -\> Show the code in the report.

-   `eval: true` -\> Run the code.

-   `warning: false` -\> Hide annoying package loading messages.

-   `label: name_of_chunk` -\> Names your chunk, this is a good practice to help trace errors in rendering.

Try it (remember to set this an an R code):

Run this chunk. Notice that the warning message is hidden in the output because we set warning: false.

```         
#| warning: false
#| label: load_packs
library(tidyverse)
print("Tidyverse loaded without warnings!")

```

## **Scoping: The Two Worlds**

One of the most common frustrations for beginners is: "It works in my console, but it fails when I Render!"

This happens because of Scoping. You are working in two different environments:

1.  **The Global Environment (Interactive Mode):**

    -   This is what you see in the "Environment" pane (top right).

    -   It contains variables you created during your current session.

    -   **The Trap:** If you create a variable `x <- 5` in the Console but forget to type it in your script, your code runs fine *now* because R finds `x` in the Global Environment.

2.  **The Document Environment (Render Mode):**

    -   When you click **Render**, Quarto starts a **brand new, empty R session**.

    -   It knows **nothing** about your Global Environment.

    -   It runs your code chunks from top to bottom. If `x` isn't defined in the script, the Render fails.

**Try it:**

1.  Create a code chunk in your Quarto document.

2.  Assign the value 1 to a variable `a` inside the chunk: `a <- 1`.

3.  Run the chunk by clicking the green "Play" button.

4.  Look at the **Environment** pane (top right). You should see `a` listed there. This is the **Global Environment**.

5.  Now, create a second code chunk below the first one.

6.  Inside this new chunk, try to print `a`: simply type `a`.

7.  Run this second chunk. It works! This is because both chunks are running in your current interactive session.

8.  **Now, delete the first chunk (where you defined `a`) but keep the second chunk.**

9.  Click the **Render** button.

10. **Result:** The render will fail with an error: `object 'a' not found`. This proves that the Render process starts fresh and doesn't know about the `a` in your Global Environment!

# **Part 2: The Backpack (Lists)**

## **Why Lists?**

-   **Vector:** A single tube of DNA. (All items must be the same type).

-   **Dataframe:** A 96-well plate. (Rows and Columns).

-   **List:** A **Backpack**.

A list can hold *anything*. It can hold a dataframe in the first pocket, a vector in the second pocket, and a single string in the third.

**Why do we care?** In Genomics (Bioconductor), complex data is stored in lists. A single object might hold the raw counts, the sample metadata, and the gene annotations all at once.

## **Creating a List**

Let's pack a backpack.

```{r}         
#| lebel: bp1
# 1. A numeric vector
my_counts <- c(100, 200, 50)

# 2. A dataframe
my_metadata <- data.frame(ID = c("A", "B"), Cond = c("WT", "Mutant"))

# 3. Create the list
my_backpack <- list(
  RawData = my_counts,
  SampleInfo = my_metadata,
  ExperimentName = "Yeast_Trial_1"
)

# Look at the structure
str(my_backpack)

```

## **Accessing List Elements**

We use the double bracket `[[ ]]` or the dollar sign `$` to reach into a pocket.

```{r bp2}         
# Grab the Experiment Name
my_backpack$ExperimentName

# Grab the Dataframe (Pocket 2)
my_backpack[[2]]

# Grab the 1st number inside the Vector inside the List (Inception!)
my_backpack$RawData[1]

```

:::{.callout-note}

Note that the \$ operator works here because we named every "pocket" in the "backpack" - or every element of the list. In unnamed lists the operator will not allow indexing of different elements.

:::

# **Part 3: The Assignment**

**Mission Brief**

You are creating a reproducible report for a Yeast RNA-Seq experiment.

You will practice:

1.  **Project Management:** Setting up the file structure.

2.  **Lists:** Storing experiment metadata.

3.  **Tidyverse:** Cleaning messy sequencing data.

4.  **Reporting:** Using Quarto chunk options.

**The Data**

A raw sequencing metrics file (raw_sequencing_data.csv) is available for download from Brightspace.

The data contains sequencing run information for Wild Type and Mutant yeast strains, but it is messy (failed runs, missing data).

## **Task 1: Project Setup**

1.  Create a new R Project named: `lastname_day2_genomics`.

2.  Create a `data` folder inside your project.

3.  Place `raw_sequencing_data.csv` in `data/`.

4.  Create a new Quarto document named: `analysis_report.qmd`.

5. Keep only the YAML header and remove the rest of the content. 

## **Task 2: The Setup Chunk (Quarto Logic)**

In your `analysis_report.qmd`:

1.  Edit the YAML to include a Table of Contents.

2.  Create a setup chunk that loads `tidyverse`, remember to name your code chunks.

3.  **Crucial:** Set the chunk options so that `tidyverse` loading messages/warnings are **HIDDEN** in the final HTML. (Hint: `warning: false`).

4.  **Formatting:** For the following tasks, use Secondary Headers (`## Task X: Title`) to clearly separate each part of your analysis.

## **Task 3: Numerical Vectors**

Let's reinforce some Day 1 concepts before moving on.

1.  Start a new code chunk.

2.  Use the `seq()` function to create a numerical vector of all the **even** numbers between 1 and 500.

3.  Assign this vector to a variable called `even`.

4.  Check that `even` is indeed a numerical vector (hint `is.numeric()` or `class()`).

5.  Subset the `even` vector to keep elements smaller than 250 and assign them to a variable named `sub_even`. (Hint: Remember that you access vectors with `[]` and inside the brackets you must explicitly refer to `even` again for the comparison, e.g., `even[even < 250]`).

## **Task 4: Character vectors and string operations**

1.  The following code uses the `sample()` function to create a random vector of 250 elements holding the characters A, T, C, G (the 4 nucleotides making up DNA) and assigns it to a variable named `DNA`. Before copying the code to your document and creating the vector, note 3 things:

    1.  `c()` inside `sample()` shows you the order of execution of chained functions in R. First, a vector of 4 elements is created by `c()`, and then a vector of 250 elements is randomly sampled from it by the function `sample()`. This would be the equivalent of `DNA <- c(...) %>% sample(...)`. You may ask, how is sampling a vector of 4 elements 250 times possible?

    2.  If we think of the vector created by `c()` as a hat, then `replace = TRUE` returns every element back to the hat once it is sampled, so it can be sampled again.

    3.  Random processes are not reproducible. If you run this code again, you will get a different sequence. In the following weeks, we will cover how to make random code reproducible, but if you’re curious, read about the function `set.seed()`.

```{r str1}         
#| eval: false
DNA <- sample(c("A","T","C","G"), 250, replace = TRUE)

```

2.  Write code that uses a pipe to create the `DNA` vector. This time, name the vector `DNA2` and use the native pipe `|>` instead of the tidyverse pipe `%>%`.

3.  Use the `==` operator to ask R if `DNA` and `DNA2` are identical. Are they? (Hint: See note 3 above).

4.  Subset the first ten elements of the vector `DNA` and assign them to the vector `DNA_sub` (Hint: `[]`).

5.  Copy the following code to turn `DNA` into a vector of one 250-character string element. The function `paste()` pastes vectors together; `collapse = ""` turns characters into a single string.

```{r str2}         
#| eval: false
DNA_string <- paste(DNA, collapse = "")

```

6.  Subset the first 150 characters of `DNA_string`, and assign them to the variable `DNA_string_sub`. (Hint: Remember Day 1? You cannot use `[]` on a string to get letters. You need the function `substr()`).

7.  Use the `nchar()` function to check that `DNA_string_sub` indeed has 150 characters.

## **Task 5: Matrices and Dataframes Refresher**

1.  Take the `DNA` vector you created in Task 4 and turn it into a matrix with 25 rows and 10 columns. Assign it to a variable named `DNA_matrix`.

2.  Extract the element found in the 15th row and 7th column of `DNA_matrix`.

3.  Create a dataframe named `df_refresher`. It should have two columns:

    -   `DNA_Seq`: containing the `DNA` vector.

    -   `Even_Num`: containing the `even` vector from Task 3.

    -   *Note:* Both vectors must be of length 250 for this to work.

4.  Use the `$` operator to access the `Even_Num` column and calculate its mean.

5.  Use square brackets `[]` to subset the dataframe and retrieve the row where the `Even_Num` value is exactly 100. (Hint: `df[df$Column == Value, ]`). **Do not** use `filter()`.

## **Task 6: The Backpack (Lists)**

Before importing data, create a List object named experiment_metadata in R.

It must contain three slots:

1.  `Exp_ID`: A string "EXP-2025-Yeast".

2.  `Date`: A string with today's date.

3.  `Conditions`: A character vector `c("WildType", "Mutant")`.

*Print this list in your report.*

## **Task 7: Data Wrangling (Tidyverse)**

1.  Import `raw_sequencing_data.csv` into a dataframe `raw_data`.

2.  Create `clean_data` using a pipe `%>%`:

    -   **Filter** to keep only rows where `Run_Quality` is "Pass".

    -   **Filter** to remove rows where `Read_Count` is `NA` (Hint: `!is.na(Read_Count)`).

    -   **Mutate** a new column `Reads_Million` (Read_Count / 1,000,000).

    -   **Select** only `SampleID`, `Condition`, `Reads_Million`, and `GC_Percent`.

## **Task 8: The Challenge (Logic Check)**

The PI wants to know the **Mean GC Content** for the "WildType" condition.

1.  Filter `clean_data` to "WildType" only.

2.  Pull the `GC_Percent` column.

3.  Calculate the mean().

    Report this number in your text.

## **Task 9: Visualization**

Create a boxplot of `Reads_Million` vs `Condition`.

-   Add a title using the `experiment_metadata$Exp_ID` list object you created in Task 6.

-   *Hint:* `ggtitle(experiment_metadata$Exp_ID)`

## **Submission**

1.  Render your report to HTML.

2.  Verify that the plot title appears and no warnings are shown.

3.  Submit the HTML file
