---
title: "The basics of transcriptomics pt3"
subtitle: "Differential Gene Experession Analysis and Visualization"
author: "Dr. Ido Hatam"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-copy: true
    self-contained: true
---

```{=html}
<style type="text/css">
body{ font-size: 18px; font-family: Arial, Helvetica, sans-serif; line-height: 30px; }
td { font-size: 14px; }
ul { font-size: 18px; }
h1.title { font-size: 38px; color: Black; }
h1 { font-size: 28px; color: Black; }
h2 { font-size: 22px; color: Black; }
h3 { font-size: 18px; font-family: Arial, Helvetica, sans-serif; color: Black; }
code.r{ font-size: 14px; }
.tocify { border: none; }
pre { font-size: 14px; }
</style>
```

# **The Road to Discovery: Day 3 Overview**

Today marks the transition from "Data Processing" to "Biological Discovery." Over the last two days, you have cleaned raw reads, mapped them to the *Arabidopsis* genome, and quantified gene expression. However, those numbers are currently split between you and your partner.

**Our goals for today:**

1.  **Collaborative Merging:** We will unify your individual count matrices into a single "Master Matrix" representing the entire experiment.

2.  **Statistical Modeling:** We will use `DESeq2` to move beyond simple averages and account for biological variance.

3.  **DGE Analysis:** We will identify the specific genes that are "significantly" triggered by both virulent and avirulent pathogens.

4.  **Exploratory Visualization:** We will use Hierarchical Clustering to audit our data and verify the quality of our replicates.

# **Part 0: Computational Environment**

Before we begin, ensure you are working in a fresh Quarto document for today's tasks.

1.  Create a new file named `transcriptomics_day3_p1.qmd` (or `_p2.qmd`).

2.  Ensure your `transcriptomics_env` is active in your terminal.

3.  Load the following libraries in your setup chunk:

```{r}         
#| eval: false

# Define the required packages
pkgs <- c("tidyverse", "DESeq2", "DT", "pheatmap", "RColorBrewer", "EnhancedVolcano", "ape")

# Load all packages efficiently using lapply
lapply(pkgs, library, character.only = TRUE)

```

# **Part 1: The Great Merger**

Yesterday, you and your partner processed different subsets of the Howard et al. 2013 data. 

 - Partner 1 handled **Mock and Virulent**
 
 - Partner 2 handled **Virulent and Avirulent**
 
 - Partner 3 handled  **Mock vs. Avirulent**

### **1. Why are we merging? (The Theory of Robustness)**

In bioinformatics, merging datasets isn't just about "getting all the numbers in one place." It is about **increasing downstream analysis robustness**.

-   **Better Noise Estimation:** `DESeq2` works by estimating "Dispersion"—how much a gene's expression varies naturally across replicates. By merging your 12 samples with your partner's 6 additional samples, the software now has **18 samples** to calculate this background noise. A larger sample size leads to a much more accurate "baseline," making your final gene list significantly more reliable.

-   **Unified Normalization:** Merging ensures all 18 samples are normalized against each other simultaneously. This prevents "batch effects" where samples processed separately look different just because of the math, not the biology.

### **2. Syncing your Files**

1.  **Partners 1 and 2:** Ensure your `Partner_Specific_Counts.csv` is committed and pushed to GitHub if you haven’t done so already.

2.  **Partner 3:** **Pull** the repository to download your partner's file into your local `data/` folder.

3.  **Confirm:** Verify that both of you now have the files needed for the merge in your `data/` directory.

### **3. Merging Dataframes**

We need to combine these two tables into a single **Master Count Matrix**. Since the Gene IDs are currently stored as row names, we must move them into a column to join them correctly, however there is a challenge here.

**1. The Challenge: Circular Redundancy**
In a group of three, every condition is represented in two different files. For example, both Partner 1 and Partner 2 have the "Virulent" samples. If we simply joined all the tables together, we would have 36 columns instead of 18, and our statistical model would fail because of perfectly duplicated data.

**2. Our Approach: The "Safe-Merge" Strategy**
Instead of manually picking columns (which is prone to human error), we will use Set Logic. We will build a pipeline that:

  1. Takes your data as the "Master" base.

  2. Looks at the next partner's file.

  3. Subtracts any column names that already exist in your Master table.

  4. Joins only the unique, missing data.

This ensures that no matter how many partners you have, the final matrix is clean, unique, and contains exactly 18 biological samples.

**Now You:** Partner 3, complete the merger in your `.qmd`. Fill in the correct file paths for your own file and the file you just pulled from your partner.

```{r}         
#| eval: false

# 1. Load all files available in your group
my_counts      <- read.csv("data/PASTE_YOUR_FILENAME.csv", row.names = 1)
partner_1_counts <- read.csv("data/PASTE_PARTNER_A_FILENAME.csv", row.names = 1)
partner_2_counts <- read.csv("data/PASTE_PARTNER_B_FILENAME.csv", row.names = 1) # Optional for Trios

# 2. Start with your counts as the base Master Matrix
# We convert row names to a column named "GeneID" to enable the join
counts_master <- my_counts %>% rownames_to_column("GeneID")

# 3. Merge Partner A - The "Safe-Merge" Pipe
# any_of() checks the column names of our master table and 
# removes them from the partner's table BEFORE joining.
counts_master <- partner_A_counts %>% 
  rownames_to_column("GeneID") %>%
  select(-any_of(colnames(counts_master)), GeneID) %>%
  inner_join(counts_master, by = "GeneID")

# 4. Merge Partner B (Only if you are a group of 3)
counts_master <- partner_B_counts %>% 
  rownames_to_column("GeneID") %>%
  select(-any_of(colnames(counts_master)), GeneID) %>%
  inner_join(counts_master, by = "GeneID")

# 5. Finalize the matrix
counts_master <- counts_master %>% column_to_rownames("GeneID")

# 6. Sanity Check: 
# Regardless of group size, you should see exactly 18 columns 
# (6 Mock, 6 Virulent, 6 Avirulent)
ncol(counts_master)
```

**Export the `counts_master` object as a CSV file to the shared `data` folder.**

### **4. The Hand-off: Moving to Individual Analysis**

**Stop and Sync:** Once the merge is successful, the partner who performed the merge must **Push** the updated `.qmd` and the final `counts_master` CSV to GitHub. The other partner(s) must **Pull**.

**From this point forward, the collaboration pauses.** While you will use the same dataset, you will each perform the statistical analysis in your own `.qmd` file to ensure you both master the logic.

# **Part 2: The Experimental Design Matrix**

Software like `DESeq2` requires a `colData` table to know which sample belongs to which group.

### **1. Creating the colData**

```{r}         
#| eval: false

# Create a metadata frame
# Ensure the levels are set so 'Mock' is the reference (the baseline)
colData <- data.frame(
  row.names = colnames(counts_master),
  Condition = factor(rep(c("Mock", "Virulent", "Avirulent"), each = 6), 
                     levels = c("Mock", "Virulent", "Avirulent")),
  Time = factor(rep(rep(c("1h", "6h", "12h"), each = 2), 3))
)
```

### **2. Task: Metadata Expansion**

Professional metadata often includes technical details about how the sequencing was performed. Even though the original study used "Paired-End" reads, we chose to simplify our pipeline on Day 2 by only aligning the **R1** files. This effectively makes our current data **single-read**.

**Now You:** Add a new column to your `colData` dataframe named `type`.

1.  Use the `rep()` function and `factor()` to fill this column with the string `"single-read"` for every row.

2.  Print your updated `colData` to verify it now has three columns: `Condition`, `Time`, and `type`.

**Critical Check:** Compare `rownames(colData)` and `colnames(counts_master)`. They **must** be identical in order and spelling for the next step to work.

**Hint:** Instead of manually checking 18 names, use the `all()` function and the `==` operator to verify they match: `all(rownames(colData) == colnames(counts_master))`. If this returns `TRUE`, you are safe to proceed.

# **Part 3: Introduction to DESeq2**

### **1. Negative Binomial Distribution**

RNA-seq data consists of counts. Unlike height or weight (Normal distribution), counts follow a **Negative Binomial Distribution**. `DESeq2` is built to handle this, adjusting for sequencing depth and the high variance found in genes with low counts.

### **2. Constructing the DESeq Object**

```{r}         
#| eval: false
# Create the dds object
dds <- DESeqDataSetFromMatrix(countData = round(counts_master),
                              colData = colData,
                              design = ~ Condition)
```

# **Part 4: Exploratory Data Analysis: Hierarchical Clustering**

Before we hunt for specific differentially expressed genes, we must audit our data. We need to answer a whole question: **Are our biological replicates actually similar to each other?**

### **1. Theory: The Distance of Expression**

Hierarchical clustering calculates the "distance" between samples based on their entire gene expression profile.

-   Samples that are biologically similar (e.g., two Mock replicates) should have a small distance and cluster together.

-   Samples that are biologically different (e.g., Mock vs. Virulent) should have a large distance and appear on different branches of the "tree."

### **2. Why Transform the Data?**

We cannot cluster raw counts. In raw data, a few extremely high-expression genes would dominate the calculation, making all samples look identical. We use the **Variance Stabilizing Transformation (VST)** to "level the playing field" so that mid-range and low-range genes contribute equally to the distance calculation.

### **3. Task: Building the Dendrogram**

```{r}         
#| eval: false

# 1. Transform the data to stabilize variance
vsd <- varianceStabilizingTransformation(dds, blind = FALSE)
vsd_mat <- assay(vsd) # Extract the matrix of transformed values

# 2. Calculate the distance between samples
# We transpose the matrix with t() because dist() calculates distance between rows
sample_dist <- dist(t(vsd_mat))

# 3. Perform Hierarchical Clustering
clusters <- hclust(sample_dist)

# 4. Plot the results using the ape package for a cleaner phylogenetic look
# We convert the hclust object to a phylo object for professional plotting.
# We map 'Condition' to 'tip.label' so the tree is easy to interpret.
plot.phylo(as.phylo(clusters), 
           tip.label = colData$Condition, 
           type = "p", 
           edge.col = "blue", 
           edge.width = 2,
           no.margin = TRUE)
```

**Now You:** Run the clustering code in your `.qmd`.

1.  Look at the dendrogram. Do the replicates for the same condition (e.g., the two `Mock` samples at 1h) cluster together on the same small branches?

2.  Do you see a clear split between the `Mock` group and the infected groups?

3.  If a sample is "out of place" (e.g., a Virulent sample sitting in the middle of the Mock branch), what might that suggest about the experiment?

# **Part 5: The Main Event - Differential Expression**

### **1. Running the Pipeline**

The `DESeq()` function is the primary statistical engine of the package. It executes a three-part workflow to ensure your results are statistically sound:

1.  **Estimation of Size Factors**: This step performs normalization.

2.  **Estimation of Dispersion**: `DESeq2` examines replicates to estimate the "noise" (dispersion) for every gene.

3.  **The Wald Test**: This is the hypothesis test to determine statistical significance.

```{r}         
#| eval: false
#| label: dds1
# Assign the output to a new object to keep your original dds safe
dds_results <- DESeq(dds)
```

### **2. Extracting Specific Results**

We use **Contrasts** to compare specific groups. Note the use of the `alpha` parameter below. In statistics, **Alpha** here represents your target **False Discovery Rate (FDR)**.

While a standard p-value measures the probability of a false positive for a *single* test, the FDR measures the *proportion* of false positives among all your significant results. Typically, researchers use 0.05 (5%). However, because we are using a **subsampled dataset**, we will use an alpha of **0.2 (20%)**. This means we accept that 20% of our "significant" genes might be false positives, in exchange for having enough signal to actually see patterns.

```{r}         
#| eval: false

# Extract results for Avirulent vs. Mock
res_avr_mock <- results(dds_results, contrast = c("Condition", "Avirulent", "Mock"), alpha = 0.2)

# Print the first few rows to see the table
print(res_avr_mock)
```

#### **Understanding the Results Table**

When you print the results, you will see several columns:

-   **baseMean**: The average of normalized counts across all samples.

-   **log2FoldChange**: The magnitude of change. A value of 1 means expression doubled ($2^1 = 2$).
 
  - Remember the rules of log if $log_{b}x = y$ then $b^y = x$

-   **lfcSE**: The standard error of the fold change.

-   **stat**: The Wald statistic used to calculate the p-value.

-   **pvalue**: The probability of a false positive for *this specific gene*.

-   **padj**: The **Adjusted P-value**. This corrects the p-value for the thousands of tests we are running. **To control the False Discovery Rate (FDR) at 20%, we filter for padj \< 0.2.**

```{r}         
#| eval: false
# See a high-level summary of the results
summary(res_avr_mock)
```

**Now You:** 1. Create a results object called `res_vir_mock` comparing **Virulent** to **Mock**.
2. Create a results object called `res_vir_avr` comparing **Virulent** to **Avirulent**.
3. Use the `summary()` function on both to check the up/down regulated counts. Which comparison shows the most "activity"?

# **Part 6: Filtering and Ranking**

### **1. Magnitude vs Significance**

Identifying biologically meaningful genes requires two filters:

1.  **Significance (`padj`):** This tells us the change is real and not noise.

    -   *p-value*: The probability that a single gene appeared significant by chance.

    -   *padj*: The probability that a gene in your **entire list** is a false positive (FDR). We will use a cutoff of **0.2** (20% FDR).

2.  **Magnitude (`log2FoldChange`):** This tells us the change is large enough to matter biologically.

    -   *Fold Change*: A simple ratio ($\frac{Group A}{Group B}$).

    -   *log2FoldChange*: A standardized scale where 0 is no change, 1 is a 2-fold increase, and -1 is a 50% decrease. We will use a threshold of **1**.

### **2. Automation with Custom Functions**

When you have multiple comparisons, it is better to write a **function** to process them.

**Important Note on `alpha` vs. Filtering:** You might wonder why we still need to filter for `padj < 0.2` if we already set `alpha = 0.2` in our `results()` call. Setting `alpha` only informs the internal "independent filtering" algorithm to optimize power for that threshold—it does **not** subset the table. The results object still contains all 30,000 genes, so we must manually filter to isolate our "Significant" list.

```{r}         
#| eval: false

# Function to filter and count DE genes
filter_and_count <- function(res_obj, comparison_name, fc_threshold = 2) {
  # 1. Remove rows with NAs in padj
  res_filtered <- res_obj[!is.na(res_obj$padj) & !is.na(res_obj$log2FoldChange), ]
  
  # 2. Apply filters: |log2FC| >= log2(fc_threshold) and padj < 0.2
  # Note: 0.2 matches the alpha we set in our results() calls
  sig_genes <- res_filtered[abs(res_filtered$log2FoldChange) >= log2(fc_threshold) & res_filtered$padj < 0.2, ]
  
  # 3. Count up and down regulated
  up_regulated <- sum(sig_genes$log2FoldChange > 0)
  down_regulated <- sum(sig_genes$log2FoldChange < 0)
  
  # Return as a clean data frame
  return(data.frame(
    Comparison = comparison_name,
    Up_regulated = up_regulated,
    Down_regulated = down_regulated
  ))
}
```

### **3. Generating the Global Summary**

Now we can use `rbind()` to stack the results into one master table.

```{r}         
#| eval: false

results_summary <- rbind(
  filter_and_count(res_vir_mock, "Vir vs Mock"),
  filter_and_count(res_avr_mock, "Avr vs Mock"),
  filter_and_count(res_vir_avr, "Vir vs Avr")
)

# Inspect the final summary table
print(results_summary)
```

**Now You:** 1. Run `summary(results_summary)` to see the distribution of your counts.
2. Prepare this data for plotting. Convert the data from "Wide" format to "Long" format using `pivot_longer()`.

**Task:** Complete the code below and generate a bar plot using `ggplot2`.

```{r}         
#| eval: false

# Step 1: Pivot the data to long format
plot_data <- results_summary %>%
  pivot_longer(cols = c(Up_regulated, Down_regulated), 
               names_to = "Regulation", 
               values_to = "Count") %>%
  mutate(Regulation = factor(Regulation, levels = c("Up_regulated", "Down_regulated")))

# Step 2: Create a Bar Plot
# Hint: Use aes(x = Comparison, y = Count, fill = Regulation) 
# and geom_bar(stat = "identity", position = "dodge")
```

# **Day 3 Wrap-Up**

Today, you merged your data to increase statistical power and verified your data quality using Hierarchical Clustering. You have successfully isolated the genomic signature of a pathogen attack.

**Where we are going tomorrow:**
In our final day, we will go from "Gene IDs" to "Biological Functions." We will use **EnhancedVolcano** plots and **Gene Set Enrichment (GO)** to see exactly how the plant's immune system is responding.

# **Final Submission**

### **Task: Exporting Results**

Export your filtered Virulent results as a CSV for tomorrow's visualization.

```{r}         
#| eval: false
# YOUR TASK: Write the code to save your significant virulent genes to 'data/data_yourname/sig_virulent_genes.csv'
```

### **Task: Reflection**

In your `README.md`:

1.  How does merging all 18 samples help `DESeq2` estimate "dispersion" more robustly?

2.  Why is a 2-fold change (`log2FoldChange` of 1) often the minimum threshold for biological interest?

3.  **Git Check:** Commit and push your `.qmd` and your exported CSV.
