---
title: "Course 3, Day 2: Functional Genomics"
subtitle: "Structural & Functional Annotation from theory to practice"
author: "Dr. Ido Hatam"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-copy: true
    self-contained: true
---

```{=html}
<style type="text/css">
body{ font-size: 18px; font-family: Arial, Helvetica, sans-serif; line-height: 30px; }
td { font-size: 14px; }
ul { font-size: 18px; }
h1.title { font-size: 38px; color: Black; }
h1 { font-size: 28px; color: Black; }
h2 { font-size: 22px; color: Black; }
h3 { font-size: 18px; font-family: Arial, Helvetica, sans-serif; color: Black; }
code.r{ font-size: 14px; }
.tocify { border: none; }
pre { font-size: 14px; }
</style>
```
# **Overview: The "You Do"**

Today, you transition from running a pipeline to analyzing its results. Working in pairs, you will compare the biological content of your **Polished** vs. **Unpolished** yeast assemblies.

**The Research Question:** Does polishing the genome (correcting indels) actually recover more biological information (genes/pathways), or is the raw assembly "good enough"?

# **Phase 1: Project Setup & Data Migration**

**Goal:** Establish a shared, reproducible workspace linked to `GitHub`.

### **1. GitHub Collaboration Setup**

Since you are working in pairs, you will share a single repository. This requires a specific "Push-Pull" workflow to get started.

1.  **Partner 1:**

    -   Log in to GitHub and create a new repository named `Course3_Day2_GroupX` (e.g. group 1).

    -   Go to **Settings** > **Collaborators** > **Add people** and invite **Partner 2 (and 3 if have one)**.

2.  **Partner 2 (and 3):**

    -   Accept the invitation via email or `GitHub` notifications.

3.  **All Partners:**

    -   Clone the repository to your local machine as a new RStudio Project (version control project).

4.  **The "Sync" Workflow:**

    -   **Partner 2:** Create two folders in your project: `data/` and `figures/`. Create a blank text file named `.gitkeep` **inside each of the folders** this is a placeholder file so Git tracks the folders. **Commit and Push** these changes to GitHub.

    -   **Partner 1(&3):** Click the **Pull** button in the RStudio Git tab to receive the new folders.
    
    -   **All partners:** Within data and figures create folders for your assebly (keep it short). Add a `.gitkeep' file to each of them. Pull and then push.

### **2. Individual Workflow & Naming Conventions**

To avoid "Merge Conflicts" (where Git gets confused by two people editing the same line), you will work in separate files before merging.

-   **Individual Files:** Each partner must create their own Quarto file for their specific assembly (e.g., `analysis_polished.qmd` and `analysis_unpolished.qmd`).

-   **Unique Filenames:** **This is critical.** Any file you generate (plots, filtered tables, etc.) must have a unique name identifying the assembly.

    -   *Example:* `fermentation_plot_polished.png` vs. `fermentation_plot_unpolished.png`.

-   **The Final Stitch:** At the end of the lab, **Partner 2** is responsible for creating the final master document by "stitching" the code and results from both partners into a single report.

### **3. Retrieve Your Annotation Data**

Copy your results into your new `data/` folder. You need:

-   `*.gff` (The structure)

-   `*.emapper.annotations` (The function)

::: {.callout-important}
Any file imported to `data/` and not placed in the `.gitignore` must have a unique name, otherwise GitHub might encounter some conflicts. Therfoer, make sure that your partner's gff files do not have identical names to yours. 
:::


### **4. Load Libraries**

```{r}         
#| eval: false
#| label: setup

#Loop that checks if the packages GO.db and AnnotationDbi are already installed
#If not, install them

p_check <- c("GO.db", "AnnotationDbi")

for(i in seq_along(p_check)){
  
  if(p_check[i] %in% installed.packages()[,"Package"]){
    print(paste("Package",p_check[i], "is already installed", sep = " "))
  }else{
    print(paste("installing package", p_check[i], sep = " ")) 
    BiocManager::install(p_check[i], force = TRUE)
  }
}
#Install the ggVennDiagram package
install.packages("ggVennDiagram")
#Load packages
library("tidyverse")
library("Gviz")
library("GenomicFeatures")
library("rtracklayer")
library("Biostrings")
library("GO.db")
library("AnnotationDbi")
library("ggVennDiagram")
```

# **Phase 2: Structural Visualization (Zooming In)**

**Goal:** Visualizing the gene density on your largest contigs.

### **1. Identifying the Largest Contigs**

-   **Task:** Import your GFF as a `TxDb` object using `makeTxDbFromGFF()`.

-   **Task:** Extract the sequence lengths and identify the top 5 largest contigs.

::: {.callout-tip}

### **Leads for Analysis**

-   **The Package:** The `seqlengths()` function is part of the **`GenomeInfoDb`** package. It allows you to see the total size of every scaffold in your assembly.

-   **Naming Consistency:** In `Gviz`, the `chromosome` name you provide to `plotTracks()` must match the **SeqID** column in your GFF exactly.

-   **Memory Management:** Start by zooming into a 50kb or 100kb window using the `from` and `to` arguments in `plotTracks()`.

:::

### **2. Visualizing Gene Models**

-   **Task:** Use `Gviz` to plot the **largest** contig in your assembly.

-   **Assignment Question 1:** Look at the gene density. Do the genes look intact (long, distinct blocks) or fragmented (many small, clustered boxes)? Compare this visual with your partner's assembly.

# **Phase 3: Integrating Function (The Overlay)**

**Goal:** Connect your EggNOG results to your Gviz structural map.

### **1. Importing Functional Data**

-   **Task:** Import the `.emapper.annotations` file using `read_tsv()`. Remember to skip the comment lines (`comment = "##"`).

### **2. Merging Structure and Function**

-   **Task:** Import the GFF as a `GRanges` object using `import.gff3()`.

-   **Task:** Perform a `left_join` or `match` operation to add a `symbol` column to your GRanges object, populated by the `Preferred_name` from EggNOG.

### **3. The Integrated Plot**

-   **Task:** Re-plot the largest contig using `GeneRegionTrack`, but this time set `transcriptAnnotation = "symbol"`.

-   **Assignment Question 2:** Pick **one gene** that now has a name. Search for it on the [Saccharomyces Genome Database (SGD)](https://www.yeastgenome.org/ "null"). What is its function? Does its location in a gene-dense or gene-sparse region make sense?

# **Phase 4: Global Functional Summary (Zooming Out)**

**Goal:** Summarize the biology of your organism using GO terms.

### **1. Data Wrangling: Splitting GO Terms**

-   **Task:** Use `separate_rows(GOs, sep = ",")` to create a "long" format dataframe.

::: {.callout-tip}

### **Level Up: The Bioconductor Way**

Today we are using `tidyverse` functions like `separate_rows` because they are intuitive and robust for simple text manipulation. However, as you advance in bioinformatics, you will encounter specialized packages designed specifically for hierarchical data like GO terms and KEGG pathways.

-   **`KEGGREST`**: Allows you to query the KEGG API directly from R to get the most up-to-date pathway maps.

-   **`AnnotationDbi` & `GO.db`**: The industry standards for mapping Gene Ontology IDs to their definitions and finding parent/child relationships between terms.

As you build your skills, look into these packages for more rigorous enrichment analyses!

:::

### **2. Visualizing Enrichment**

-   **Task:** Create a bar plot (`ggplot2`) showing the **Top 10 most common GO terms**.

Step 1. Converting GO numbers into readable terms

```{r}
#| eval: false
#| label: GOs
#Convert to long table
your_data_long <- your_data %>%
  separate_rows(GOs, sep = ",")

#Get the terms and the ontology
go_terms <- AnnotationDbi::select(GO.db,
                                  keys = unique(your_data_long$GOs),
                                  columns = c("TERM"),
                                  keytype = "GOID")

your_data_long <- left_join(your_data_long, go_terms,
                            by = c("GOs" = "GOID"))

```
::: {.callout-tip}

### **Hint for Plotting: What is `stat = "identity"`?**

When making bar plots with `ggplot2`, the default behavior is `stat = "count"`, which asks R to count the rows for you. However, if you have already summarized your data into a table with a "Count" column, you must use `stat = "identity"` (or `geom_col()`) to tell ggplot: *"Don't count the rows, just plot the values I gave you in the 'y' column exactly as they are."*

This is critical for stacked bar plots or any plot where you have pre-calculated the height of the bars.

```{r}
#| eval: false
#| label: Gos_bar

top_terms <- your_data_long %>%
  count(TERM, sort = TRUE) %>%
  filter(!is.na(TERM)) %>%
  slice_head(n = 10)

ggplot(top_terms, aes(x = reorder(TERM, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top 10 GO Terms",
       x = "GO Term",
       y = "Count") +
  theme_minimal()
```

:::

-   **Assignment Question 3:** Based on the top **Biological Process (BP)** GO terms identified, what are the primary biological capabilities encoded within this genome? Does this functional profile align with the metabolic and cellular requirements of *Saccharomyces cerevisiae*? In your answer, distinguish between **genomic potential** (what the organism is capable of doing based on its DNA) and **transcriptomic expression** (what the organism is actively doing in a specific environment).

# **Phase 5: Comparative Analysis (The Deep Dive)**

**Goal:** Quantify the "completeness" of your genome by comparing the Polished vs. Unpolished results.

### **1. The Bird's Eye Comparison (Global GO Profile)**

Sit with your partner and compare your **Top 10 GO Term Bar Plots** side-by-side.

-   **Assignment Question 4:** Did polishing "recover" entire functional categories?

    -   *What do we mean by functional categories?* We are looking for broad biological processes (e.g., **"Ribosome Biogenesis"**, **"Cell Wall Organization"**, or **"DNA Repair"**) that appear as major bars in the Polished assembly plot but are either missing or significantly shorter in the Unpolished plot.

    -   *Discussion:* Pick one such category and discuss why sequence-level errors (like indels) might have hidden these functions in the raw data.

### **2. Metabolic Pathway Completeness (Glycolysis & Fermentation)**

-   **Task:** Filter your annotation dataframe for descriptions containing "Glycolysis" or "Fermentation".

-   **Task:** Count the number of unique genes identified for each process.

-   **Task:** Visualize these counts! Create a simple bar plot comparing the number of genes found for each pathway in your assembly.

::: {.callout-tip}

### **Hint for Filtering**

To find specific pathways, focus your search on the **`Description`** column of your EggNOG results. Because gene descriptions can vary in capitalization (e.g., "Glycolysis" vs "glycolysis"), use `regex()` with `ignore_case = TRUE`.

```{r}
#| eval: false
#| label: emaper0
# Example search terms: "glycolysis", "fermentation"
pathway_hits <- eggnog_data %>%
  filter(str_detect(Description, regex("glycolysis", ignore_case = TRUE)))

```

:::

### **3. The Synthesis**

-   **Task** Compare the annotations achieved from polished and unpolished assemblies.
  
  -   Retrieve your partner's annotation data and create a list containing the datasets.
  
  -   Extract the shared column from each data frame
  
```{r}
#| eval: false
#| label: ven_setup
sets <- lapply(list_of_annotation_data, function(df) df$Preferred_name[df$Preferred_name!= "-"])
names(sets) <- names(list_of_annotation_data)
```

  - Create a Venn Diagram using `ggVennDiagram()`.
  
  - How do the annotations you received compare to those of your partners?

-   **Assignment Question 5:** Final Comparison Table:

    -   **Polished Assembly Gene Count (Fermentation):** _________

    -   **Unpolished Assembly Gene Count (Fermentation):** _________

    -   **Analysis:** Discuss the "Biological Cost of Noise." How do frameshift errors in Nanopore data lead to a loss of functional annotation? Think about how a single indel can turn one long, functional gene into two small "hypothetical" fragments.

# **Submission**

1.  Knit your final merged RMarkdown to HTML.

2.  Commit all changes to GitHub (.qmd, HTML, and Figures).

3.  Submit the link to your repository to your instructor.
