---
title: "Course 3, Day 1: Functional Genomics"
subtitle: "Structural & Functional Annotation"
author: "Dr. Ido Hatam"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-copy: true
    self-contained: true
---

```{=html}
<style type="text/css">
body{ font-size: 18px; font-family: Arial, Helvetica, sans-serif; line-height: 30px; }
td { font-size: 14px; }
ul { font-size: 18px; }
h1.title { font-size: 38px; color: Black; }
h1 { font-size: 28px; color: Black; }
h2 { font-size: 22px; color: Black; }
h3 { font-size: 18px; font-family: Arial, Helvetica, sans-serif; color: Black; }
code.r{ font-size: 14px; }
.tocify { border: none; }
pre { font-size: 14px; }
</style>
```
# **Pre-Lab Checklist: The Digital Mise-en-place**

Before we begin our analysis, we must ensure all data and software dependencies are in place. Please complete the following checklist immediately.

### **1. Data Management (The "Sneaker-Net")**

-   [ ] **EggNOG Database:** Locate the shared USB drive circulating in the lab. Copy the `eggnog_data` folder to your local `Documents/` folder.

-   [ ] **Migrate Assemblies:** Copy `assembly.fasta` (Unpolished) and `assembly_polished.fasta` (Polished) from your `Course_2/` folder to your current project `data/` folder.

-   [ ] **Demo Data:** Download `chr1.fasta` and `reference.gff` (S288C) from the course website to your `data/` folder.

### **2. Software & Licensing**

-   [ ] **Docker:** Verify you have the annotation container.

```{bash}
#| eval: false
#| label: fpull
docker pull nextgenusfs/funannotate:v1.8.15
```

-   [ ] **GeneMark License:** Register for the academic license (GeneMark-ES/ET) at the GeneMark website.

    -   Download the `gm_key` file.

    -   Place it in your user Home directory (`~/.gm_key`).

-   [ ] **Conda Environment:** Create the environment for functional annotation.

```{bash}
#| eval: false
#| label: emconda
conda create -n eggnog_env python=3.8
conda activate eggnog_env
conda install -c bioconda -c conda-forge eggnog-mapper
```

### **3. R Environment**

-   [ ] **R Project:** Initialize a new RStudio Project for "Course 3".

-   [ ] **Package Installation & Loading:** Execute the code chunk below. This script checks if the required genomic visualization and data wrangling packages are installed. If missing, it installs them automatically and then loads them into your session.

```{r}
#| eval: false
#| label: setup   
# List of required packages for this course
pkg_list <- c("Gviz", "GenomicFeatures", "rtracklayer", "Biostrings", "tidyverse")

# Conditional installation and loading loop
invisible(lapply(pkg_list, function(pkg) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    message(paste("Installing", pkg, "..."))
    if (pkg %in% c("Gviz", "GenomicFeatures", "rtracklayer", "Biostrings")) {
      if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
      BiocManager::install(pkg, update = FALSE, ask = FALSE)
    } else {
      install.packages(pkg, repos = "[http://cran.us.r-project.org](http://cran.us.r-project.org)")
    }
    library(pkg, character.only = TRUE)
  } else {
    message(paste(pkg, "is already installed and loaded."))
  }
}))
```

# **Phase 1: Structural Annotation (Theory & Demo)**

Genome annotation is the process of attaching biological information to sequences. It occurs in two stages:

-   **Structural Annotation:** Identifying the physical coordinates and properties of genomic features.

    -   *Primary Structure:* Identifying sequence elements like genes, regulatory regions, and repeats.

    -   *Modifications:* In modern long-read genomics, this also includes mapping base modifications (e.g., methylation, epigenomics), though today we focus on gene models.

-   **Functional Annotation:** Assigning biological information regarding the element's biochemical function and cellular location (e.g., "This gene encodes a kinase located in the mitochondria").

## **The what, why, and how?**

### **How do we "see" genomic elements?**

We rely on specialized **annotation software** to predict where features exist. These tools generally employ two main strategies:

1.  **Ab initio ("From the Beginning"):**

    -   *The Logic:* Uses mathematical models (like Hidden Markov Models) to recognize the statistical patterns of DNA.

    -   *The Signals:* It scans for Start codons (ATG), Stop codons, specific splice-site motifs, and **CpG islands** (regions prone to methylation that often mark gene promoters).

    -   *Pros/Cons:* Fast and cheap, but can generate false positives if the model isn't trained well.

2.  **Evidence-Based (Homology):**

    -   *The Logic:* We align real biological data ("clues") to the genome. If a sequence matches known biology, a feature likely exists there.

    -   *The Sources:*

        -   **Proteins:** Aligning protein sequences from a closely related species (e.g., pulling the *S. cerevisiae* proteome from RefSeq).

        -   **Transcripts:** Aligning RNA-seq data to the genome. This is the "gold standard" because it represents genes that are actively transcribed. You can use public data or your own experimental RNA-seq reads.

### **The Funannotate Pipeline**

`Funannotate` is a pipeline that integrates **both strategies** to create a high-quality **consensus**.

-   **What is a Consensus?** Imagine asking three different architects to design a house. They might disagree on where the windows go. `Funannotate` runs multiple prediction algorithms (the architects) and compares them against real data (the evidence). Where they disagree, it uses a weighted scoring system to decide the final, most likely gene structure—prioritizing physical evidence (Proteins/Transcripts) over statistical guesses.

#### **Order of operations**

When working with `funannotate`, or any other annotator, the workflow follows four general steps.

1.  **Cleaning:** Removes small, redundant contigs (think of those as "genomic dust").

2.  **Masking:** Identifies repetitive elements (transposons, simple repeats) and "masks" them (turns them to 'N' or lowercase). *Why?* To prevent *ab initio* tools from mistakenly calling a repetitive region a gene.

3.  **Predicting:**

    -   Runs *Ab initio* tools (Augustus, GeneMark).

    -   Aligns *Evidence* (Proteins/Transcripts) if provided.

4.  **Filtering:** Removes overlapping or biologically impossible predictions.

### **The Output: Anatomy of a GFF3 File**

The **General Feature Format (GFF3)** is the standard "address book" for a genome.

The 3 in GFF3 stands the 3rd iteration (or version) of the format. It was designed to improve upon GFF2 by providing a standardized way to define hierarchical relationships (using ID and Parent tags) and ensuring that metadata in Column 9 follows a strict, predictable structure.

**Visualizing a Faux GFF3:**

Here is what a GFF3 file looks like.

```         
##gff-version 3
# SeqID   Source      Type    Start   End     Score  Strand  Phase  Attributes
chrI      RefSeq      gene    1000    2500    .      +       .      ID=gene01;Name=YAL001W
chrI      RefSeq      mRNA    1000    2500    .      +       .      ID=mRNA01;Parent=gene01
chrI      RefSeq      exon    1000    1200    .      +       .      ID=exon01;Parent=mRNA01
chrI      RefSeq      CDS     1000    1200    .      +       0      ID=cds01;Parent=mRNA01
```

-   **Note:** The `Parent=` attribute links the parts together (Exons belong to mRNA, mRNA belongs to Gene).

## **Demo: Chromosome 1**

We will run a test on *S. cerevisiae* Chromosome 1. This allows us to verify the Docker pipeline and see the intermediate files (like the masked genome) in minutes.

### **Step 1: Clean the Sequence**

Annotation tools often crash on tiny, messy fragments. We use `funannotate clean` to remove duplicate contigs or very short fragments. Remember, we start by cleaning the "genomic dust".

**Understanding the Docker Command:**

-   `docker run`: Tells the docker engine to spin up a new container.

-   `-u $(id -u):$(id -g)`: **User Mapping.** Runs the container as *you* (your current user ID) instead of the "root" user. This ensures that any files created by the container can be edited or deleted by you later.

-   `-v "$PWD":/data`: **Volume Mounting.** Bridges the gap between your computer and the container. It maps your current working directory (`$PWD`) to the `/data` folder inside the container. This allows the software inside to "see" your input files and write output files back to your computer.

- `--platform linux/amd64`: Specifies which platform the container should run on. Since our computers are running M1/M4 chips, we set it to amd64

-   `nextgenusfs/funannotate...`: The name of the software image to run.

-   `funannotate clean ...`: The actual scientific command we are executing.

```{bash}
#| eval: false
#| label: clean
  docker run -u $(id -u):$(id -g) \
    --rm -v "$PWD":/data -w /data \
    --platform linux/amd64 \
    nextgenusfs/funannotate:v1.8.15 \
    funannotate clean \
    -i data/chr1.fasta \
    -o data/chr1_clean.fasta
```

### **Step 2: Mask Repeats**

Before predicting genes, we must hide repetitive elements so the software doesn't get confused.

```{bash}
#| eval: false
#| label: mask
docker run -u $(id -u):$(id -g) \
    --rm -v "$PWD":/data -w /data \
    --platform linux/amd64 \
    nextgenusfs/funannotate:v1.8.15 \
    funannotate mask \
    -i data/chr1_clean.fasta \
    -o data/chr1_masked.fasta \
    --cpus 10
```

### **Step 3: Predict Genes** ~5 minutes

Now we run the prediction on the **masked** genome. Note the `--augustus_species` flag; this loads pre-trained parameters for Yeast. Using a pre-trained model significantly reduces the computational expense of the prediction algorithms. Without this flag, the software must perform de novo training to learn the species' specific gene structures—a process that is remarkably computationally intensive and can extend runtimes from minutes to several hours.

Something about how funannotate predict combines other tools and how we dont have all of them set up right now.

**Understanding the Docker Command:**

-   `-s "Saccharomyces cerevisiae`: sets the species name (for naming purposes) 

-   `--augustus_species ...`: sets the pretrained model for the Augustus gene prediction tool

```{bash}
#| eval: false
#| label: predict
docker run -u $(id -u):$(id -g) \
    --rm -v "$PWD":/data -w /data \
    --platform linux/amd64 \
    nextgenusfs/funannotate:v1.8.15 \
    funannotate predict \
    -i data/chr1_masked.fasta \
    -o annotation_demo \
    -s "Saccharomyces cerevisiae" \
    --augustus_species saccharomyces_cerevisiae_S288C \
    --cpus 10
```

::: callout-note
### **Expected Output Files**

Upon successful completion, the `annotation_demo/predict_results` folder will contain several key files:

-   **`*.gff3`**: The primary structural annotation file containing coordinates for all predicted features. This is what you will use for visualization in R.

-   **`*.proteins.fa`**: A FASTA file containing the amino acid sequences of all predicted protein-coding genes. This is the **required input** for functional annotation (EggNOG-mapper).

-   **`*.transcripts.fa`**: A FASTA file containing the nucleotide sequences of the predicted transcripts (including UTRs where predicted).

-   **`*.tbl`**: A feature table file used for NCBI submission.

-   **`*.stats.txt`**: A summary report containing statistics such as gene count, average gene length, and number of exons.
:::

# **Phase 2: Visualization - "Seeing the Biology"**

*Goal: Demystify the GFF file by turning text into a map.*

## **Why visualize the GFF?**

As we've seen, a GFF file is essentially a massive, text-based spreadsheet of coordinates. While this is perfect for a computer, it is nearly impossible for a human to interpret.

Visualizing the GFF file against your genome serves as your first **Critical Sanity Check**:

1.  **Validation of Structure:** Do the "boxes" (exons) make sense? Are they clustered into gene models with reasonable gaps (introns), or do they look like random noise scattered across the contig?

2.  **Verifying Orientation:** You can see which strand a gene is on. If your assembly is reversed, the annotations will show you how the biological "flow" is oriented.

3.  **Evaluating ab initio performance:** *Ab initio* tools are statistical guessers. By plotting the results, you can see if the software predicted multiple "mini-genes" in a region that should logically be one single gene, or if it missed a region that looks like it should contain a feature.

4.  **Reference Comparison:** By looking at the Reference GFF alongside our own demo results, we can start to "spot the difference" and understand how genome assembly quality directly impacts the software's ability to find genes.

In this "We Do" exercise, we will visualize the annotations you just generated for Chromosome 1.

**Pro-Tip:** Annotation software often renames the input sequences to a standardized format (e.g., `NC_001133.9` might become `chr1`). Before plotting, we check the SeqID in the GFF.

```{r}
#| eval: false
#| label: gviz
# 1. Prepare the Data
# Path to the GFF generated in the "I Do" step
# Funannotate names the output based on the species
gff_path <- "annotation_demo/predict_results/Saccharomyces_cerevisiae.gff3"

# Create a Transcript Database (TxDb) from your new GFF
txdb_demo <- makeTxDbFromGFF(gff_path, format = "gff3")

# 2. Check Sequence IDs
# This prints the chromosome/contig names found in your file
seqlevels(txdb_demo)

# 3. Define the Track
# Use the ID found in step 2
# If our ID is an NCBI name (e.g. NC_00113.9) rather than a UCSC name (e.g. chr1), we set the ucscChromosomeNames flag to false
options(ucscChromosomeNames=FALSE)

demo_track <- GeneRegionTrack(txdb_demo, 
                             chromosome = seqlevels(txdb_demo)[1],
                             name = "Predicted Genes",
                             transcriptAnnotation = "gene",
                             background.title = "darkgreen")

# 4. Add a Ruler
axis_track <- GenomeAxisTrack()

# 5. Plot a specific region
# We plot the first 50kb of our prediction
plotTracks(list(axis_track, demo_track),
           from = 1, 
           to = 50000,
           chromosome = seqlevels(txdb_demo)[1],
           main = "Funannotate Predictions: Chromosome 1")
```

# **Phase 3: Functional Annotation (Theory)**

*Goal: Assign biological meaning to the predicted shapes.*

## **Theory Block**

### **1. The Problem: The "Hypothetical Protein"**

Structural annotation (Phase 1) gives us **coordinates** (e.g., "Gene X is located at Chr1:1000-2000"), but it doesn't tell us **what** Gene X actually does. Without functional annotation, our genome is filled with "Hypothetical Proteins"—sequences that possess all the structural hallmarks of a gene (Start/Stop codons, Open Reading Frames) but have no assigned biological role. This is the "dark matter" of genomics. Our goal in this phase is to illuminate it.

### **2. The Solution: Orthology vs. Homology**

To predict the function of a hypothetical protein, we usually compare our unknown gene to databases of known genes. There are two ways to do this, and the distinction is critical:

-   **Homology (e.g., BLAST):** This searches for sequences that share a common ancestor. However, homology is broad—it includes **Paralogs** (genes duplicated within a genome that often evolve *new* functions). If you BLAST your gene and it matches a paralog with a different function, your annotation will be incorrect.

-   **Orthology (e.g., EggNOG):** This is more precise. Orthologs are genes separated by a **speciation event**. They typically retain the **same function** across different species.

**Why EggNOG-mapper?** We use EggNOG-mapper because it uses pre-computed evolutionary trees to place your protein into a specific **Orthologous Group**. This allows for highly accurate functional transfer because it distinguishes between "looks like" (homology) and "descended from the same functional ancestor" (orthology).

### **3. The Vocabulary of Function**

We don't just want a gene name; we want standardized, computer-readable data that allows us to compare biology across the world.

-   **GO (Gene Ontology):** Biology's "Controlled Vocabulary." Instead of one scientist saying "sugar transporter" and another saying "glucose carrier," we use the ID `GO:0005355`. This standardization is crucial for computational analysis.

    -   *Molecular Function:* What the protein does (e.g., "Kinase activity").

    -   *Biological Process:* The larger goal it contributes to (e.g., "Glycolysis").

    -   *Cellular Component:* Where it lives (e.g., "Mitochondria").

-   **KEGG (Kyoto Encyclopedia of Genes and Genomes):** The **Map**. Knowing a gene is a "Kinase" is useful, but knowing it is the "rate-limiting step in Glycolysis" (Map00010) tells you exactly how it fits into the cell's metabolic wiring diagram.

    -   *Example:* You find a gene assigned to **K00128**. This maps to the **Glycolysis / Gluconeogenesis pathway (map00010)**, allowing you to confirm that your yeast strain possesses the enzymatic machinery for ethanol production.

-   **EC (Enzyme Commission) Numbers:** The **Chemistry**. These numerical codes (e.g., `1.1.1.1`) describe the exact chemical reaction an enzyme catalyzes, regardless of the protein sequence.

    -   **First digit:** The major class of the reaction (e.g., 1 = Oxidoreductases, 2 = Transferases, 3 = Hydrolases, etc.).

    -   **Second digit:** The sub-class (often the type of bond or group acted upon).

    -   **Third digit:** The sub-sub-class (often the specific acceptor or co-factor).

    -   **Fourth digit:** The specific serial number for the enzyme.

    -   *Example:* **Alcohol dehydrogenase (EC 1.1.1.1)**. The first '1' tells us it is an Oxidoreductase; the second '1' indicates it acts on the CH-OH group of donors; the third '1' shows it uses NAD+ or NADP+ as an acceptor; and the final '1' identifies it specifically as alcohol dehydrogenase.

-   **COG (Clusters of Orthologous Groups):** A broad classification of protein ancestry, grouping proteins into general functional categories (e.g., "Amino acid transport and metabolism").

    -   *Example:* You observe that 15% of your predicted genes fall into **Category \[J\]: Translation, ribosomal structure and biogenesis**. This high percentage is a hallmark of rapidly growing cells like yeast during exponential phase, where protein synthesis machinery is a priority.

### **4. The Input & Output**

-   **The Input:** To run functional annotation, we need the amino acid sequences of our predicted genes.

    -   *Recall Phase 1:* In the demo, `funannotate` generated a file named `annotation_demo/predict_results/Saccharomyces_cerevisiae.proteins.fa`. This file contains the translation of every gene model we found. This is exactly what we feed into EggNOG.

-   **The Output:** EggNOG-mapper will produce an `.emapper.annotations` file. This is a comprehensive, tab-delimited table that links your specific Gene IDs (e.g., `FUN_0001`) to all the vocabularies listed above (GO, KEGG, EC, Gene Names).

# **Phase 4: Functional Annotation in Action (The "We Do")**

*Goal: Generate biological data and connect it to our visualization.*

We will now run the Functional Annotation pipeline on our Chromosome 1 demo data. This is a shorter version of what you will do overnight for the full genome.

### **Step 1: Run EggNOG-mapper (Terminal)**

We use `emapper.py` to classify our proteins.

-   **Input:** The protein FASTA file from Phase 1 (`*.proteins.fa`).

-   **Database:** The `eggnog_data` folder you copied earlier.

-   **Method:** We use `-m diamond` for fast alignment.

-   **Output** We use `--decorate_gff yes` and `excel` to get outputs in gff and excel formats

**Action:** Copy and run this command in your Terminal.

```{bash}
#| eval: false
#| label: enmap1
# Activate the environment
eval "$(conda shell.bash hook)"
conda activate eggnog_env

# Run EggNOG
# Adjust EGGNOG_DB_PATH to match where you put the folder (e.g., /Users/student/Documents/eggnog_data)
EGGNOG_DB_PATH="/Users/$(whoami)/Documents/eggnog_data"

emapper.py \
-i annotation_demo/predict_results/Saccharomyces_cerevisiae.proteins.fa \
--output chr1_functional \
--output_dir annotation_demo \
--data_dir $EGGNOG_DB_PATH \
--decorate_gff yes \
--excel \
-m diamond \
--cpu 10
```

### **Step 2: Import & Clean Data (R)**

Once the run finishes (approx. 5-10 mins), we import the results into R. The output file (`.emapper.annotations`) contains many comment lines that we need to handle.

```{r}
#| eval: false
#| label: tidymap
library(tidyverse)

# Define the path to your new file
eggnog_file <- "annotation_demo/chr1_functional.emapper.annotations"

# Import the data
# We use comment = "##" to skip the top metadata lines
# The header line starts with #query, so we rename it immediately
eggnog_data <- read_tsv(eggnog_file, comment = "##", show_col_types = FALSE) %>%
  rename(query = `#query`)

# Inspect the data
view(eggnog_data)
```

### **Step 3: Mining for Biology (R)**

Hypothesis: Since this is *Saccharomyces cerevisiae* (Baker's Yeast), we expect to find genes related to **fermentation**. Let's check our annotation for evidence.

```{r}
#| eval: false
#| label: plot_ferment
# Filter for genes with "fermentation" in their description
fermentation_genes <- eggnog_data %>%
  filter(str_detect(Description, regex("fermentation", ignore_case = TRUE)))

# View the hits
print(paste("Found", nrow(fermentation_genes), "fermentation-related genes."))

# Plot the top Gene Names found
ggplot(fermentation_genes, aes(x = Preferred_name)) +
  geom_bar(fill = "steelblue") +
  theme_minimal() +
  labs(title = "Fermentation Genes Identified on Chr 1",
       x = "Gene Name",
       y = "Count") +
  coord_flip()
```

No hits. Let's consider what could have led to this result. Let's get an overview of the annotations we got from Eggnog-Mapper. One good way to do th

```{r}
#| eval: false
#| label: plot_COG



```

### **Step 4: Updating the Map (R)**

Now for the final integration. We will overlay these biological names onto our structural map.

-   **The Logic:** The GFF file contains the structure. The EggNOG table contains the names. We link them using the **Transcript ID** (`query` column in EggNOG matches `ID` in GFF).

```{r}
#| eval: false
#| label: gviz2

library(Gviz)
library(rtracklayer)

# 1. Import GFF as a GRanges object (preserves all metadata)
gff_gr <- import.gff3("annotation_demo/predict_results/Saccharomyces_cerevisiae.gff3")

# 2. Filter the GFF for only predicted mRNA (what we annotated) 

gff_gr <- gff_gr[gff_gr$type == "mRNA"]

# 3. Match IDs to Names
# We look for where the GFF IDs match the EggNOG Query IDs
# We create a new column called 'symbol' in our genomic ranges
match_idx <- match(gff_gr$ID, eggnog_data$query)
gff_gr$symbol <- eggnog_data$Preferred_name[match_idx]

# 4. Fill in Missing Names
# If a gene has no name (NA), we keep the original ID so it doesn't disappear
gff_gr$symbol[is.na(gff_gr$symbol)] <- gff_gr$ID[is.na(gff_gr$symbol)]

# 4. Create the Track with Names
# Note: We use transcriptAnnotation = "symbol" to tell Gviz to use our new column
func_track <- GeneRegionTrack(gff_gr, 
                              chromosome = seqlevels(gff_gr)[1],
                              name = "Functional Annotations",
                              transcriptAnnotation = "symbol",
                              background.title = "darkblue",
                              fill = "lightblue")

# 6. Plot
plotTracks(list(axis_track, func_track),
           from = 1, 
           to = 50000,
           chromosome = seqlevels(gff_gr)[1],
           main = "Annotated Genome: Structure + Function")
```

# **Phase 5: The Production Launch (The "We Do")**

*Goal: Write the pipeline script and launch the overnight job.*

Since the full genome takes ~1 hour to predict and significant time to annotate, we will automate the process with a shell script.

### **1. Create the Script**

Start a text file in RStudio's text editor and name it `run_pipeline.sh`.
Copy the following code into your file and save it. Just add the path to your assembly, the path to the eggnog database and the name of your eggnog conda environment.

```{bash}
#| eval: false
#| label: the_script
#!/bin/bash
set -e  # Exit immediately if a command exits with a non-zero status

echo "Starting the pipeline at"
date

# Variables
IMAGE="nextgenusfs/funannotate:v1.8.15"
PLATFORM="--platform linux/amd64"
CPUS=8

# ----------------------------
# Check that input was provided
# ----------------------------
if [ -z "$1" ]; then
    echo "Usage: $0 <input_fasta>"
    exit 1
fi

INPUT="$1"

CLEAN="pipeclean.fasta"
SORTED="pipesorted.fasta"
MASKED="pipemasked.fasta"
OUTDIR="pipefunannotate"
SPECIES_NAME="Saccharomyces_cerevisiae"
AUG_SPECIES="saccharomyces_cerevisiae_S288C"

EGGNOG_ENVIRONMENT="eggnog-env"
EGGNOG_DB_PATH="/Users/mtaillefer00/eggnog_db"

echo "Step 1: Cleaning assembly..."
docker run --rm -v "$PWD":/data -w /data $PLATFORM $IMAGE \
    funannotate clean -i $INPUT -o $CLEAN

echo "Step 2: Sorting assembly..."
docker run --rm -v "$PWD":/data -w /data $PLATFORM $IMAGE \
    funannotate sort -i $CLEAN -o $SORTED

echo "Step 3: Masking assembly..."
docker run --rm -v "$PWD":/data -w /data $PLATFORM $IMAGE \
    funannotate mask -i $SORTED -o $MASKED --cpus $CPUS

echo "Step 4: Running gene prediction..."
docker run --rm -v "$PWD":/data -w /data $PLATFORM $IMAGE \
    funannotate predict \
    -i $MASKED \
    -o /data/$OUTDIR \
    -s $SPECIES_NAME \
    --augustus_species $AUG_SPECIES \
    --cpus $CPUS

echo "Step 5: Annotating predicted genes..."
# Initialize Conda (Required for scripts)
eval "$(conda shell.bash hook)"
conda activate $EGGNOG_ENVIRONMENT

# Run emapper (Pointing to the local DB path)
emapper.py \
  -i $OUTDIR/predict_results/*.proteins.fa \
  --output yeast_functional \
  --data_dir $EGGNOG_DB_PATH \
  --decorate_gff yes \
  --excel \
  -m diamond \
  --cpu $CPUS

echo "Pipeline Complete. Check 'yeast_functional.emapper.annotations'."
date
```

### **2. Launch the Job**

1.  Make the script executable:

```{bash}
#| eval: false
#| label: chmod
chmod +x run_pipeline.sh
```

2.  Launch the script and redirect output to a log file:

`2>&1` tells the computer to redirect Standard Error (file descriptor 2) to the same location as Standard Output (file descriptor 1). This ensures that both the normal progress updates and any potential error messages are captured together within the pipeline.log file rather than cluttering your terminal screen.

```{bash}
#| eval: false
#| label: run_script
./run_pipeline.sh path/to/assembly.fasta > pipeline.log 2>&1
```

Tomorrow we will check that the run didn't fail and that all expected files were outputted. 