---
title: "Week 2, Day 4: Assembly & Mapping Logic - Practice"
subtitle: "De Novo Assembly and Structural Validation- Yeast Genome Edition"
author: "Dr. Ido Hatam"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-copy: true
    self-contained: true
---

```{=html}
<style type="text/css">
body{ font-size: 18px; font-family: Arial, Helvetica, sans-serif; line-height: 30px; }
td { font-size: 14px; }
ul { font-size: 18px; }
h1.title { font-size: 38px; color: Black; }
h1 { font-size: 28px; color: Black; }
h2 { font-size: 22px; color: Black; }
h3 { font-size: 18px; font-family: Arial, Helvetica, sans-serif; color: Black; }
code.r{ font-size: 14px; }
.tocify { border: none; }
pre { font-size: 14px; }
</style>
```

# **The Scenario: The Assembly Arm-Race**

Today, we complete the lifecycle of genome assembly. You are no longer just following a tutorial; you are conducting a comparative experiment. You and your partner will test two different assembly strategies to see which yields a more biologically accurate representation of the *Saccharomyces cerevisiae* genome.

# **Part 0: Logistics & Shared Infrastructure**

Before starting, you must establish your collaborative environment.

## **1. The Shared Repository**

1.  **Partner A:** Go to your `Week2_Yeast_Project` repository on GitHub.

2.  **Settings:** Go to **Settings** -> **Collaborators** -> **Add people**. Add **Partner B** & **Partner C**.

3.  **Partner B/C:** Accept the invitation via email or GitHub notifications.

4.  **Both:** Initialize your **RStudio Project** linked to this repository.

    -   *How to:* Go to **File** -> **New Project** -> **Version Control** -> **Git**.

    -   Paste the repository URL.

    -   Once the project opens, click **Pull** in the Git pane to ensure you are synced.

## **2. The Workflow Strategy (Independent Work, Joint Report)**

To prevent merge conflicts while you are both working, we will use a "Modular" approach:

-   **Partner A:** Create a file named `arm_a_baseline.qmd`.

-   **Partner B:** Create a file named `arm_b_iterative.qmd`.

-   **Partner C:** Create a file named `arm_c_polished.qmd`.

-   **The Final Merge:** At the end of the session, **Partner B** (as the Lead Editor) will merge the content from all arms into one cohesive document: `Final_Assembly_Report.qmd`.

Make sure that the code and answers to all the steps are in your `.qmd` files. 

## **3. Git Hygiene: Protecting Your Repository**

Large biological data files should **never** be pushed to GitHub. However, **result reports** (HTML, text summaries) from tools like QUAST and BUSCO should be included.

1.  Open your `.gitignore` file from the Files pane.

2.  **Ignore Data:** Add the following to ensure your raw and processed data stays local:

* `.gitignore`

* `fastq` files

* `bam` files

* `.gz` files

Any other file that's >50mb

3.  **Verification:** Check your Git pane. If you still see large files with a status icon, they are being tracked and must be removed.

## **4. Reference Download and Preparation (Programmatic Renaming)**

If you haven't yet download the S. cerevisiae reference genome from `RefSeq`.
Same as with the assembly, we are doing it via R so it is easy to chain to other components of the workflow. 

```{r}
#| label: reference-dl
#| eval: false
#Address
URL <-"https://ftp.ncbi.nlm.nih.gov/genomes/refseq/fungi/Saccharomyces_cerevisiae/latest_assembly_versions/GCF_000146045.2_R64/GCF_000146045.2_R64_genomic.fna.gz"

#Destenatoin file (output)

dest_file <- "GCF_000146045.2_R64_genomic.test.fna.gz"

#Download the file with wget
download.file(URL, method = "wget", destfile = dest_file)

```


The NCBI `RefSeq` reference genome uses IDs like `NC_001133.9`, but genome browsers (and `Gviz`) prefer UCSC conventions like `chrI` or `chr1`. To ensure our entire workflow is reproducible, we will include a code chunk to "fix" our reference headers.

:::{.callout-note}
We didn't cover if-else statements in our introductory course, but we are using one here to check if our file is complete.
Note the syntax, they can be useful once you start developing more elaborate workflows. 
:::

```{r}         
#| label: reference-prep
#| eval: false

# 1. Uncompress and Read the raw NCBI file as simple text
con <- gzfile("GCF_000146045.2_R64_genomic.fna.gz", "rt")
ygen <- readLines(con)

close(con)  # <-- this is for closing the connection to the gz file, saves on resources. 



# 2. Define the new UCSC-style headers (chrI through chrXVI + chrM)
new_headers <- c(paste0(">chr", as.roman(1:16)), ">chrM")

# 3. Find and Replace
header_lines <- grep(">", ygen)
#Look at the header for chromosome 1
header_lines[1]

# Safety Check: Yeast has 16 nuclear chromosomes + 1 mitochondrial
if(length(header_lines) == 17) {
  ygen[header_lines] <- new_headers
  writeLines(ygen, "S288C_UCSC.fna")
  message("SUCCESS: Reference renamed to UCSC standards.")
} else {
  stop("Error: Chromosome count mismatch! Check your input file.")
}

```

# **Part 1: The Experimental Design**

We are testing the effect of **iterative internal polishing** on assembly contiguity and accuracy.

-   **Arm A (The Baseline):** `Flye` with 3 internal iterations + Scaffolding + `Medaka` polishing.

-   **Arm B (The Iterative Arm):** `Flye` with 5 internal iterations + Scaffolding + `Medaka` polishing.

-   **Arm C (The Polished Arm)** `Flye` with 5 internal iterations + Scaffolding + 2 Rounds of `Medaka` polishing.


**Partner A will execute Arm A. Partner B will execute Arm B. Partner C will execute Arm C.**

# **Part 2: The Reconstruction (Bash)**

## **Step 1: Assembly with Flye & Log Audit**

1.  **Run Flye:** Construct your command using `yeast_filtered.fastq.gz`. Ensure you enable `--scaffold` and set the correct number of `--iterations`.

2.  **Audit the Polishing:** Open `flye.log`. Record the number of "errors corrected" for each iteration.

3.  **Extract Final Stats:** Copy the summary lines (Size, Contigs, N50) into your `.qmd`. Compare your total assembly size to the expected \~12.1 Mb genome.

## **Step 2: Polishing with Medaka**

Take your final `assembly.fasta` from Flye and run one round of `medaka_consensus`.

-   **Note on Model:** Our data is from a legacy chemistry. **Do not set the `-m` flag**; let Medaka auto-detect the correct model for you.

# **Part 3: Technical & Biological Audit (Docker)**

Now, evaluate your results. **Important:** Remember to commit and push these reports to GitHub so your partner can see them!

## **1. QUAST (Technical Quality)**

Run QUAST using your new `S288C_UCSC.fna` as the reference.

-   **Task:** Once finished, **push the results folder to GitHub**.

-   **Metric focus:** Extract the **N50**, **Total Length**, and number of **Misassemblies**.

Compare the two assemblies, do you see any differences in coverage, accuracy, contiguity?

## **2. BUSCO (Biological Completeness)**

Run BUSCO using the `saccharomycetaceae_odb12` database.

-   **Task:** Once finished, **push the summary report to GitHub**.

* What were your BUSCO scores for each assembly? Is one more complete than the other?

# **Part 4: Structural Validation (R & Gviz)**

## **Step 1: High-Efficiency Alignment & Indexing**

We map our assembly back to the reference using a high-efficiency pipe to save disk space.

### **The Table of Contents: The .bai file**

Once the BAM file is sorted, we must create an **index** file (`.bai`). This is the "Index Card" that tells a computer exactly where each read is located, enabling the "Google Maps" style scrolling in **Gviz** or **IGV**.

```{bash}
#| label: minimap
#| eval: false
# Map and Sort in one step
# minimap2 -a (output SAM) | samtools sort (input from pipe)
minimap2 -a -x asm5 data/reference/S288C_UCSC.fna medaka_output/consensus.fasta | samtools sort -o assembly_mapped.sorted.bam -

# Create the index
samtools index assembly_mapped.sorted.bam

```

## **Step 1.5: Global Coverage Audit**

```{bash}
#| label: coverage
#| eval: false
# Run the coverage report
samtools coverage assembly_mapped.sorted.bam -o assembly_cov.txt

```

```{r}         
#| label: coverage-audit
#| eval: false
# Read the coverage report
assembly_cov <- read.table("assembly_cov.txt", sep = "\t", header = TRUE, comment.char = "")
print(assembly_cov)

```

* Are there any differences in coverage of chromosomes between each of the assemblies?

## **Step 2: Programmatic Discovery & Selection**

Before we visualize, we will calculate the mapped ranges for **every chromosome**. However, for our visual audit, we will select one specific "Target Chromosome" to focus on.

**Selection Criteria for Visualization:**

1.  Identify chromosomes with a **length \> 500,000 bp**.

2.  From that group, choose the one with the **highest number of mapped contigs** (the `mapped` column in `idxstatsBam`).

```{r}         
#| label: mapping-discovery
#| eval: false
library(Rsamtools)
library(GenomicAlignments)

# 1. Audit the Stats: Find chromosomes > 500kb with high density
stats <- idxstatsBam("assembly_mapped.sorted.bam")
print(stats)

# 2. Extract ranges for the entire genome
bam_file <- "assembly_mapped.sorted.bam"
alignment_ranges <- bam_file |> readGAlignments() |> GRanges()

# Look at the ranges for ALL chromosomes to see your global success
ranges(alignment_ranges)

```

## **Step 3: The Custom Visual Audit**

Use `Gviz` to plot your assembly against the reference.

::: {.callout-important}

### **The "Why" Behind the Code: Messy Biology vs. Clean Plots**

You might notice that your assembly is missing huge chunks of the genome (low coverage). **This is intentional!** To let you run this lab on your own computer in minutes, we gave you a tiny **'snapshot'** of the data.

In addition, this data comes from an older generation of sequencing technology (Nanopore R7). It has a higher error rate, leading to more fragmented contigs.

Lastly, Yeast genomes are plastic. When the software (`minimap2`) tries to force our 'square peg' contig into the 'round hole' of the reference genome, it realizes the ends don't fit.

**What this means for your project:**
When `minimap2` aligns these messy reads, it uses **Soft Clipping**. Think of it like a sticker that is too long for the page: the aligner sticks down the part that matches, and lets the rest 'hang off the edge.'

If we try to plot the BAM file directly, `Gviz` sees those hanging edges, gets confused about coordinates, and often crashes.

**The Fix:** We load the data into R as a `GRanges` object first (Step 2). This acts like a **pair of scissors**—it trims off the unaligned 'hanging edges' and keeps only the matching parts. This allows us to visualize exactly where our DNA matches the reference without the software panicking.
:::

**Your Task:**

1.  Use the `alignment_ranges` object you created in Step 2 to build your `AlignmentsTrack`.

2.  Refer back to your `ranges()` output from Step 2. Pick the chromosome you identified in the selection step.

3.  Manually enter the chromosome name and the coordinates into `plotTracks()`.

```{r}         
#| label: visual-audit
#| eval: false
library(Gviz)

# 1. Create the Data Track using the GRanges object (The "Scissors" approach)
align_track <- AlignmentsTrack(alignment_ranges, name = "Assembly Alignments", fill = "darkblue")

# 2. Create the Ruler
axis_track <- GenomeAxisTrack()

# 3. Plot
# Replace the chromosome name and numbers with the ones you discovered!
plotTracks(list(axis_track, align_track), 
           from = "[ENTER_START_NUMBER]", 
           to = "[ENTER_END_NUMBER]", 
           chromosome = "[ENTER_CHR_NAME]")

```

Did you end up selecting different chromosomes? What does your inspection reveal? Was one approach more successful in alligning more accurately along the reference genome?

# **Part 5: The Grand Synthesis (Partner B Task)**

Partner B will merge the modular reports into `Final_Assembly_Report.qmd`.

-   **Head-to-Head Table:** Compare polishing errors, N50, Misassemblies, BUSCO scores, and Chromosome Coverage.

-   **Stacked Visualization:** Stack the `Gviz` tracks for Arm A, Arm B and Arm C on top of each other against the reference.

# **Final Assignment**

-   **The Log Audit:** Did errors corrected decrease significantly by the 5th iteration?

-   **Completeness vs. Accuracy:** Did iterative polishing (Arm B & C) increase biological completeness (BUSCO) or accuracy (QUAST)?

-   **The Coverage Check:** Look at `assembly_cov`. What % of the chromosomes did you cover?

-   **The Size Check:** How does your size compare to \~12.1 Mb?

-   **The Visual Audit:** Did the stacked tracks reveal any structural differences resolved by Arm B or C?

**Submission:** Render to HTML. Push all code and results. Submit the shared link.
