---
title: "Week 2, Day 2: Omics & Data Hygiene"
subtitle: "From Signal to Sequence: The Garbage In, Garbage Out The Practice"
author: "Dr. Ido Hatam"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-copy: true
    self-contained: true
---

```{=html}
<style type="text/css">
body{ font-size: 18px; font-family: Arial, Helvetica, sans-serif; line-height: 30px; }
td { font-size: 14px; }
ul { font-size: 18px; }
h1.title { font-size: 38px; color: Black; }
h1 { font-size: 28px; color: Black; }
h2 { font-size: 22px; color: Black; }
h3 { font-size: 18px; font-family: Arial, Helvetica, sans-serif; color: Black; }
code.r{ font-size: 14px; }
.tocify { border: none; }
pre { font-size: 14px; }
</style>
```
# **Introduction to You Do**

Yesterday, we walked through the theory and practice of sequencing data hygiene. Today, you take the wheel. You will independently clean two distinct datasets: a messy short-read dataset from a microbial community analysis and a long-read yeast genome that will serve as your primary project for the next two weeks.

# **Part 1: Short-Read Hygiene (Mothur MiSeq Data)**

This dataset comes from the famous **Mothur MiSeq SOP** (Standard Operating Procedure), developed by Pat Schloss's lab at the University of Michigan. It is one of the most widely used tutorials in microbial ecology.

**The Context:** The data comes from a study analyzing the gut microbiome of mice. Specifically, it looks at how the microbial community structure remains stable after weaning (transitioning from milk to solid food). It is a classic example of **16S rRNA amplicon sequencing** performed on an Illumina MiSeq instrument. Because it is real biological data, it contains all the messy artifacts—adapters, quality drops, and errors—that you will encounter in your own research. In addition, this data-set was sequenced *A WHILE* ago, which means that the phred scores and mean per read quality score will not be as nice as with modern illumina sequencing kits. This will be a great learning experience, understanding what messy data looks like. **Note:** As always with example data in this course, this is a **subset** of the original dataset, ensuring you can process it quickly on your own computers.

## **Step 1: Project Setup**

1.  **Initialize:** Create a **new** GitHub repository named `Week2_ShortRead_QC`.

2.  **Clone:** Start a new RStudio Project via Version Control linked to this repo.

3.  **Document:** Create a new Quarto document `ShortRead_Hygiene.qmd`.

## **Step 2: Data Acquisition**

In bioinformatics, R can acts not just as a statistical tool, but as a facilitator of workflows. Instead of jumping back and forth between the terminal and RStudio, we can execute file system commands directly from R. This approach improves reproducibility because every step—from creating folders to downloading raw data—is documented in your script and executed via a single console.

First, let's download and extract the data using R functions instead of bash commands.

```{r dl_and_unzip}
#| eval: false
# Create data directory
dir.create("data/raw_reads", recursive = TRUE)

#The adress for data download
URL <- "https://mothur.s3.us-east-2.amazonaws.com/wiki/miseqsopdata.zip"

#download the file
download.file(url = URL, destfile = "data/raw_reads/miseqsopdata.zip", method = "wget")

#check that the zip file was downloaded
list.files("data/raw_reads/")
# Unzip it
unzip ("data/raw_reads/miseqsopdata.zip", exdir = "data/raw_reads/")

```

Next, lets check that the `.fastq` files are there. It is important to make sure that both the forward and the reverse sequencing files are there for each sample. Files with `R1` are the forward and those with `R2` are the reverse.

```{r listfiles}
#| eval: false

list.files("data/raw_reads/MiSeq_SOP")

```

Note that `.fastq` are not the only files in this folder. We will need to conciser that when specifying input folders later on.

## **Step 3: The Audit (Rqc)**

You already have the tools. Now use them to generate a complete diagnostic profile.

1.  **Load Libraries:** Ensure `Rqc`, `ShortRead`, `QuasR` and `ggplot2` are loaded.

2.  **Run Rqc:** Point the `rqc()` function to the `MiSeq_SOP` directory and assign the output to an object (e.g., `qc_result`).

3.  **Generate Diagnostic Plots:** Create and save the following 4 plots (just like Day 1). **Recall Challenge:** Use your notes to find the correct functions!

    -   **Average Read Quality:** Is the overall library healthy?

        -   *Task:* Generate this plot and save it as `average_quality.png`.

    -   **Per Cycle Quality:** Do you see the 3' drop?

        -   *Task:* Generate this plot and save it as `per_cycle_quality.png`.

    -   **Per Cycle Base Content:** Is the composition stable?

        -   *Task:* Generate this plot and save it as `base_content.png`.

    -   **Read Frequency:** Are there duplicates?

        -   *Task:* Generate this plot and save it as `read_frequency.png`.

4.  **Generate the HTML Report:** **Recall Challenge:** Using your notes from Day 1, write a code chunk that generates the full interactive HTML report.

    -   *Hint:* Don't forget to set the `knitr` options (`eval = TRUE`, `echo = FALSE`, etc.) to ensure the report renders successfully in the non-interactive Quarto environment!

    -   Save the report to a clean output directory (e.g., `Rqc_Report`).

## **Step 4: The Industry Standard (FastQC & MultiQC)**

Now replicate the industry standard workflow using the terminal and R aggregation.

1.  **Terminal:** Run `fastqc` on the `MiSeq_SOP` directory. Save outputs to a new folder `fastqc_results`.

2.  **R Aggregation:** Use `qc_aggregate()` from the `fastqcr` package to summarize the results.

3.  **Summary:** Use `qc_report()` to generate the MultiQC-style summary.

## **Step 5: The Cleanup (Trimming with QuasR)**

Today we will learn how to trim our short reads. Remember what we are looking for from our QC audit:

1.  **3' Quality Drop:** We need to remove the low-quality bases at the end of the read.

2.  **Reads that are too short:** After trimming, if a read is too short to be useful, we discard it.

We will use the `QuasR` package, which you installed in Day 1's setup.

### **The Learning Phase: Trial Run**

Before we touch our massive MiSeq dataset, let's practice on the small example dataset from the `ShortRead` package (the one we used yesterday).

**Your Task:** Run the code below to see how `preprocessReads()` works. It will trim the example file and save a clean version.

```{r}
#| label: filt_short
#| eval: false

# 1. Locate the example file
example_file <- system.file(package="ShortRead", "extdata/E-MTAB-1147", "ERR127302_1_subset.fastq.gz")

# 2. Define where the output should go
clean_example <- "clean_example.fastq.gz"

# 3. Run the Trimmer
# truncationQuality = 20: Chop off bases from the end if quality < 20
# minReadLength = 50: Throw away the read if it ends up shorter than 50bp
preprocessReads(filename = example_file,
                outputFilename = clean_example,
                truncationQuality = 20,
                minReadLength = 50)
```

Make sure to list only the `fastq` files. These reads are larger than the shorRead example dataset, therefore tream the last 50bp and filter out reads smaller than 200bp.

### **The "You Do" Challenge: Trimming Your Data**

Now, apply this logic to your **Mothur MiSeq Data**.

1.  **Define Inputs:** Create a variable `fq_list` that lists all FASTQ files.

    -   *Tip:* Use pattern matching with regular expressions (`\\.fastq$`) to ensure you only select the correct files. Note the double backslash `\\` as an escape character.

```{r}
#| label: qst1
#| eval: false
# List only the fastq files with the correct path
fq_list <- paste("data/raw_reads/MiSeq_SOP/", 
                 list.files("data/raw_reads/MiSeq_SOP", pattern = "\\.fastq$"), 
                 sep = "")
```

2.  **Define Outputs:** Create a corresponding list of output filenames assigned to `out_list`.

```{r quast2}
#| eval: false
out_list <- paste("data/processed_reads/",
                  list.files("data/raw_reads/MiSeq_SOP", pattern="\\.fastq$"),
                  "_processed.fastq", sep="")
```

::: callout-important
Do not forget to use `dir.create("data/processed_reads")` prior to trimming!
:::

3.  **Execute & Analyze:** Run `preprocessReads()`. Assign the function output to a variable named `survived` to catch the log data.

    -   Check the `class()` of `survived`.

    -   **The Wrangling Challenge:** * Transform `survived` so rows become columns using `t()`.

        -   Convert to a dataframe using `as.data.frame()`.

        -   Use the native R pipe `|>` to chain these.

        -   Use `mutate()` to calculate a new column `percentSurvived` (totalPassed / totalSequences * 100).

        -   Calculate the `mean()` read survival rate across all samples.

    -   **The Display:** Install/load the `DT` package. Use `DT::datatable()` to output your transformed `survived` dataframe as an interactive table in your final HTML document.

4.  **Verify:** Run the entire `Rqc` report on the `data/processed_reads` folder to prove your data is now clean!

# **Part 2: Long-Read Hygiene (Yeast Genome Project)**

This dataset will be your companion for Weeks 2 and 3. Treat it with care.

## **Step 1: Project Setup (Separate Repo)**

1.  **Initialize:** Create a **new** GitHub repository named `Week2_Yeast_Project`.

2.  **Clone:** Start a new RStudio Project.

3.  **Why separate?** This repository will grow into a full genome assembly and annotation project. We want to keep it distinct from the short-read practice.

## **Step 2: Data Acquisition**

Just as we did with the short-read data, we will use R as our facilitator to acquire the long-read dataset. By keeping our acquisition steps in R, we ensure the entire pipeline—from raw signal to clean sequences—is reproducible and documented in a single place.

```{r lreads}         
#| eval: false
# 1. Create the data directory
dir.create("data/raw_reads", recursive = TRUE)

# 2. Define the download URL
yeast_url <- "ftp://ftp.sra.ebi.ac.uk/vol1/fastq/ERR153/001/ERR1539001/ERR1539001.fastq.gz"

# 3. Download the file
# method="wget" is used for robustness with larger sequencing files
download.file(url = yeast_url, 
              destfile = "data/raw_reads/ERR1539001.fastq.gz", 
              method = "wget")

# 4. Verify the file is present
list.files("data/raw_reads/")
```

## **Step 3: The Audit & Filter (Replication)**

Your task is to replicate the "Best Practice" workflow you learned yesterday.

1.  **Audit:** Run `sequali` on the raw data.

2.  **Prepare Output Directory:** Use R as a facilitator to create a directory named `data/clean_reads`.

```{r opfold}         
#| eval: false
dir.create("data/clean_reads")
```

3.  **Filter:** Use `filtlong` (`--min_length 1000` and either `--keep_percent 90` or `--min_mean_q 7`).

    -   **Recall:** Remember how we used the pipe (`>`) in the terminal yesterday to save the output of `filtlong` into a compressed `.gz` file? Apply that same logic here to save your clean reads into your new folder.

4.  **Verify:** Run `sequali` on the output.

**Deliverable:**
A clean, filtered FASTQ file named `yeast_filtered.fastq.gz` in your `data/clean_reads` folder.

# **Final Submission**

1.  **Short-Read Repo:** **Render** your `.qmd` to HTML. Commit and push your `.qmd` file, the rendered `.html` file, your diagnostic plots, and the generated Rqc HTML report.

2.  **Long-Read Repo:** **Render** your `.qmd` to HTML. Commit and push your `.qmd` file, the rendered `.html` file, and all `sequali` reports.

3.  **Submit:** Send links to BOTH repositories to your instructor.
