---
title: "Week 1, Day 3: The Reproducible Environment"
subtitle: "Git, Conda, and Docker"
author: "Dr. Ido Hatam"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format: 
  html:
    toc: true
    toc-depth: 3
    theme: 
      light: sandstone
      dark: darkly
    code-copy: true
    self-contained: true
---

```{=html}
<style type="text/css">
.quarto-color-scheme-toggle::before {
  content: " Dark Mode";
  margin-right: 0.3em;
}
.quarto-color-scheme-toggle {
  font-size: 1.2em !important;
  padding: 0.5em !important;
  border: 1px solid currentColor !important;
  border-radius: 4px !important;
}
/* Make the icons bigger */
.quarto-color-scheme-toggle i {
  font-size: 1.3em !important;
}
.quarto-color-scheme-toggle:hover {
  background-color: rgba(128, 128, 128, 0.2);
}
</style>
```

# **Before We Begin**

While we discuss the philosophy of reproducibility, let's get your environment ready in the background.

Please:

1.  Start an R project for today, give it a descriptive name.

2.  Start a `.R` script file, save it into a scripts folder and name it computational_environment_setup

3.  Copy the following code to `computational_environment_setup.R` and run it.

```{r conda_setup}
#| eval: false
#| 
# 1. Install the course setup package from GitHub
if (!require("devtools")){ 
  install.packages("devtools")}

# 2. Install the course setup package from GitHub

devtools::install_github("idohatam/Biol4315setup")

# 2. Initialize the environment check and conda installation
Biol4315setup::conda_whats_installed()
```

The `Biol4315setup::conda_whats_installed()` function will:

-   Install `Miniconda` (if not already installed) and ensure all necessary configurations are set.

-   Install essential third-party tools for upcoming labs.

-   Install required R packages.

It might take a few minutes, so let it run while we talk!

# **1. The Philosophy: The Crisis of Reproducibility**

You wrote a script. It runs perfectly on your laptop. You made sure the pathing is not absolute and did not use fixed values specific to your machine. You email it to your collaborator. **It crashes.**

Why?

1.  They have a different version of R or other software needed.

2.  They don't have the right packages installed (R or otherwise), often referred to as dependencies.

3.  They use a different OS (e.g. Mac vs Windows) or a computational environment.

Case in Point:

The Biol4315setup package you just ran worked beautifully here in the lab. Why? Because I designed it specifically for these Apple Silicon computers. If you tried to run this exact same code on an old Windows laptop at home, it would likely fail because the underlying binaries it tries to fetch are different.

This is a **Hardware Dependency**. Your code isn't wrong, but your environment is incompatible.

Today, we solve this "It works on my machine" crisis using the **Holy Trinity of Reproducibility**:

1.  **Git:** Tracks changes to your code (Time Machine).

2.  **Conda:** Manages your software libraries (The Tool Shed).

3.  **Docker:** Freezes your entire operating system (The Shipping Container).

# **2. The Terminal: Your New Best Friend**

To control these powerful tools, we must leave the comfort of the R Console and enter the **Terminal**.

### **The RStudio Terminal Tab**

Look next to your Console tab. Click Terminal.

This connects you directly to your computer's operating system (Shell).

::: {.callout-note}

### Micro-Challenge 1: Navigation

In the Terminal tab, type the following commands and hit Enter:

1.  `pwd` (Print Working Directory) - Where am I?

2.  `ls` (List) - What files are here?

3.  `mkdir test_folder` (Make Directory) - Create a new folder.

4.  `ls` - Verify the new folder exists.

5.  `rmdir test_folder` - Delete it.

:::

# **3. Version Control: Git & GitHub**

Before we set things up, let's clarify what we are actually doing. Beginners often use "Git" and "GitHub" interchangeably, but they are very different things.

### **What is Version Control?**

Imagine writing a paper. You save paper_v1.docx. Then your PI gives feedback, so you save paper_v2_edits.docx. Then paper_final.docx. Then paper_final_REALLY_FINAL.docx.

This is messy and confusing.

**Version Control** systems track every single change you make to a file automatically. You don't save new files; you save "snapshots" of the project in time. You can travel back to any previous snapshot instantly. It is a **Time Machine** for your project.

**Why is this useful?** It prevents disaster by allowing you to:

1.  **Reverse Code Breaks:** If an update to R or a package breaks your code, you can revert to the exact version that worked yesterday.

2.  **Fearless Experimentation:** You can try a new, risky analysis method. If it fails, you just delete the "branch" and your original code remains untouched.

3.  **Prevent File Clutter:** No more `analysis_final_v2_fixed.R`. You have one file, `analysis.R`, with a complete history of every change ever made to it.

4.  **Recover Deleted Work:** If you delete a chunk of code and realize two weeks later you actually needed it, it is still in the history.

### **Git vs. GitHub**

-   **Git** is the software that runs on your laptop. It creates the time machine. It works even if you have no internet connection. Git has its own syntax and is run via the terminal.

-   **GitHub** is a website (a cloud service). It hosts a copy of your Git project so you can share it, back it up, or collaborate with others. A copy of a Git project hosted on GitHub is called a repository (or repo for short).

**Analogy:** Git is like **writing** in a Word doc on your laptop. GitHub is like **saving** that doc to Google Drive or Dropbox so others can see it.

R’s `usethis` package helps manage the connection between Git and GitHub, and RStudio has a very convenient interface to manage that connection. It makes integrating Git and GitHub into project work very easy and intuitive.

## Setting up Git and GitHub:

### Step 0: GitHub account:

Make sure you signed up to a GitHub account, if you haven’t please do so now. Make sure to note the email and password used for the account as you’ll need those to connect Git and GitHub.

### **Step 1: Configuration (The Easy Way)**

We will use the R package `usethis` to configure Git. This saves us from memorizing cryptic terminal commands.

Run this **in your R Console** (not Terminal):

```{r gitconfig}
#| eval: false
library(usethis)

# 1. Tell Git who you are
# IMPORTANT: Use the exact SAME email address and username you used for your GitHub account.
use_git_config(user.name = "Your Name", user.email = "your_email@example.com")

# 2. Verify it worked
git_sitrep()
```

### **Step 2: Connecting Git with GitHub and RStudio (GitHub First Approach)**

The easiest way to connect Git, GitHub, and RStudio is the **GitHub First** approach. We start by creating the repo in the cloud.

1.  On your GitHub page locate and click the **Repositories** tab.

2.  Click **New**, name it `reproducible_env1_yourinitials` (e.g., `reproducible_env1_ih`).

3.  **Crucial:** Check the box **Add a README file**. This initializes the repo so you can clone it immediately.

4.  Click **Create repository**.

To securely connect Git and GitHub you’ll need a Personal Access Token (PAT).

```{r gtoken}
#| eval: false
# This will open a browser window to GitHub
create_github_token()
```

1.  In the browser, give the token a description (e.g., "RStudio Lab").

2.  Make sure the token is valid for 90 days (after that you’ll have to renew the token).

3.  Scroll down and click **Generate Token**.

4.  **COPY THE TOKEN IMMEDIATELY.** You will never see it again.

Back in RStudio:

```{r gitcreds}
#| eval: false
library(gitcreds)
# This function securely stores your GitHub Personal Access Token (PAT)
# in your computer's credential store so you don't have to type it every time.
gitcreds_set()
# Paste your token when asked.
```

### **Step 3: Cloning (Bringing it Home)**

**Cloning** is the process of creating a local copy of a remote repository on your own computer. It downloads the entire history of the project from GitHub and establishes a link so you can push and pull changes.

You have a repo on GitHub. You have RStudio configured. Now you need to link them.

1.  Go back to your new GitHub repository page.

2.  Click the green **Code** button.

3.  Copy the URL (HTTPS).

4.  In RStudio: Go to `File` -\> `New Project`.

5.  Select **Version Control** -\> **Git**.

6.  Paste the URL into **Repository URL**.

7.  Click **Create Project**.

*RStudio will now download (Clone) your GitHub repo to your computer and open it as a new Project.*

::: callout-note
What just happened?

Under the hood, RStudio is running the command `git clone \<URL\>`. This is one of the most fundamental Git commands. You will likely encounter `git clone` frequently in your bioinformatics career, especially when installing software or downloading pipelines directly from GitHub.
:::

### **Step 4: The Workflow**

Now that they are connected, you use the **Git Pane** in RStudio (top right) to sync them.

1.  **Pull:** Always pull first to make sure that the local and cloud repositories are in sync. This is especially important when you're not the only one working on the repo or when working from multiple computers (e.g., home and lab). (Blue Arrow)

2.  **Commit:** Save a snapshot of your changes locally. (Checkboxes -\> Commit)

3.  **Push:** Send your snapshot to the cloud. (Green Arrow)


::: {.callout-important}

## **Micro-Challenge 2: Testing the Connection**

Let's verify your Git-GitHub-RStudio connection is actually working before we move on to bigger tasks.

### **Phase A: The README and the First Push**

1.  Open the `README.md` file in your project.

2.  Add a line: "This repository contains my Week 1 Day 3 computational environment setup and notes."

3.  **Commit your change:**

    -   In the Git Pane, check the box for `README.md` (Stage).

    -   Click **Commit**.

    -   **The Message Matters:** A good commit comment tells your future self *what* you did. Use "Updated README with project description" rather than just "edit".

4.  Click **Push** (Green Arrow).

5.  Refresh your GitHub page in the browser to confirm the change is there.

### **Phase B: Organizing Scripts**

1.  In your RStudio Files pane, create a new folder called `scripts`.

2.  Move the `computational_environment_setup.R` file into this new `scripts/` folder.

3.  **Commit and Push** this change with a useful message.

### **Phase C: Starting the Notebook**

1.  Create a new Quarto document (`.qmd`) named `day3_reproducibility_notes.qmd`.

2.  Set up the YAML header with a Title and Table of Contents. Delete the generic example text.

3.  Document your R setup: Create a header `## R Environment Setup`.

4.  Write a sentence about what you did, and add the setup code into an R chunk (use `#| eval: false`).

5.  Commit and Push.

:::

# **4. Package Management: Conda**

In Genomics, we use tools written in Python, C++, and Perl. RStudio (and R) cannot install these alone because they require their own specialized compilers and libraries. We need **Conda**.

### **What is Conda?**

Conda is a cross-platform package and **environment** manager. While R uses `install.packages()` to manage R-specific libraries, Conda manages *everything else*. It is the industry standard for bioinformatics because it handles complex dependencies automatically.

We specifically use **Miniconda**, which is a "light" version of the software. While the full "Anaconda" version comes with hundreds of pre-installed data science packages (most of which you'll never use), Miniconda provides just the bare essentials. This keeps your system clean and allows you to install only exactly what you need.

The core power of Conda lies in its ability to create isolated **environments**.

::: callout-tip
The "No Admin" Bonus

One major advantage of Conda is that it installs everything inside your own "User" folder. You do not need Administrative Privileges to install software. This allows you to bypass locked-down IT departments and manage your own scientific tools!
:::

### **What is an Environment?**

Imagine you have two projects:

-   Project A needs `Tool A` version 1.0.

-   Project B needs `Tool A` version 2.0.

If you install them globally on your computer, they **might** conflict and break each other. This happens because of versioning; for example, `Tool A` version 1.0 might require Python 3.8, while `Tool A` version 2.0 requires Python 3.11. Since your computer can usually only have one version of a software as the default at any given time, installing one often overrides or breaks the other.

**Conda** lets you create separate "boxes" (environments) for each project. Within a box, you can have a specific version of Python, R, and any number of bio-tools without worrying about the rest of your computer.

### **Setting up Conda: A Step-by-Step Installation**

We are now going to use Conda to install a common Quality Control tool for Illumina sequencing called `fastqc`. We will do this by following the standard professional workflow.

::: callout-note
Why specify Illumina?

fastqc is designed for short-read sequencing data (like Illumina). If you were working with long-read data (Oxford Nanopore or PacBio), you would use different tools such as NanoPlot or PycoQC, as the error profiles and data structures are very different.
:::

#### **Step 1: Search for availability**

Before installing anything, we search to make sure the tool is available for our specific computer architecture. We also specify the `bioconda` "channel".

In the **Terminal**:

```{bash csearch}
#| eval: false
conda search -c bioconda fastqc
```

The search results will show multiple available versions. By default, Conda installs the **latest** version available. If you need a specific version, you can specify it using the `=` operator. e.g. `my_favorite_conda_package = 0.1`

#### **Step 2: Create a new Environment**

We create a clean, isolated "box" for our genomics tools.

```{bash ccreate}
#| eval: false
conda create --name genomics -y  # -y means "yes" to all questions automatically
```

#### **Step 3: Activating the Environment**

Activating means "stepping inside the box." Note that your terminal prompt will usually change to show `(genomics)` at the start.

```{bash cactivate}
#| eval: false
conda activate genomics
```

::: callout-note
How do I get out?

To exit the environment and return to your normal computer settings, simply type `conda deactivate`.
:::

#### **Step 4: Installation**

Now that we are inside our `genomics` box, we can safely install the tool.

```{bash cinstall}
#| eval: false
conda install -c bioconda fastqc -y
```

#### **Step 5: Verification (The Help Flag)**

The best way to check if a tool is correctly installed is to try and access its "Help" menu.

```{bash helpflag}
#| eval: false
fastqc -h
```

::: callout-important
## **Micro-Challenge 3: Documenting the Tool Shed**

Now that you've mastered Conda, you need to document it. This is where we practice **Polyglot** Quarto—using multiple languages in one report.

1.  In your `day3_reproducibility_notes.qmd` document, create a new section `## Conda Environment Setup`.

2.  For **every step** you just performed (Search, Create, Activate, Install, Help):

    -   For this exersize you want to create a new environement called week1 and install fastqc version 0.11.8 in it.

    -   Don't forget to deactivate the previous environment you created if it is still active (you should see base in your terminal)

    -   Create a secondary header (e.g., `### Step 1: Searching for FastQC`).

    -   Write a line explaining what that specific step does.

    -   Add a **Bash code chunk** (```` ```{bash} ````) instead of an R chunk. This tells Quarto these are terminal commands.

    -   **Crucial:** Add `#| eval: false` to every bash chunk so Quarto doesn't try to install software when you render.

3.  **Commit after every code chunk you add.**

4.  **The Goal:** Make sure your commit comments are descriptive (e.g., "Documented environment activation").

5.  Push after every commit.
:::

# **5. Containerization: Docker**

Sometimes, even Conda isn't enough. If Conda is a "Tool Shed" for specific software, Docker is an entire **Shipping Container** for the house. It freezes the **entire operating system** (Ubuntu, Linux, etc.) into a single file called an **Image**.

### **Why Docker?**

In genomics, some tools are incredibly difficult to install. They might require 50 different **dependencies** (other software packages like `blast+`) that only work on one specific version of Linux. Furthermore, each dependency might have to be a specific version, which increases the complexity. This is especially true for modern long-read quality check tools like `NanoPlot` or `PycoQC`. Instead of spending three days trying to install it, you can just download a Docker Image where the developer has already done the work for you.

### **Docker Desktop**

To use Docker, you must have **Docker Desktop** installed and **running** on your computer.

-   **The Engine:** Docker Desktop is the "engine" that runs in the background. If you close the app, your terminal commands will fail.

-   **The Dashboard:** The app has a visual dashboard where you can see which images you've downloaded and which "containers" (running instances) are active.

-   **Architecture:** Docker handles the "translation" between your Mac/Windows hardware and the Linux environment inside the container.

### **Images vs. Containers**

1.  **Image (The Recipe):** A read-only file that contains the operating system and the software.

2.  **Container (The Meal):** A running instance of an image. You can start, stop, and delete containers without affecting the original image.

::: {.callout-tip}
**The Future of Software Deployment**
Containerized environments are becoming the industry standard for how software is supplied and deployed. If you eventually work on High-Performance Computing (HPC) clusters, like the ones managed by the **Digital Research Alliance of Canada**, you will find that using containers is often the easiest (and sometimes only) way to run software that isn't already pre-installed by the system administrators.
:::

### **Pulling an Image**

In this example, we are pulling **BUSCO** (Benchmarking Universal Single-Copy Orthologs).

**What is BUSCO?**
BUSCO is a gold-standard tool in genomics used to assess the "completeness" of a genome assembly. It works by searching your data for a set of universal, single-copy genes that *should* be present in any healthy organism of a specific lineage.

**Why are we using Docker for this?**
BUSCO is **notoriously hard to install**. It has a massive web of dependencies (other software it needs to run) that are very sensitive to versions. If you try to install it manually on a university or workplace computer, you will likely hit two walls:

1.  **The Admin Wall:** You often don't have the "Administrative Privileges" required to install the underlying system libraries.

2.  **The Complexity Wall:** Even with admin rights, getting all the dependencies to "talk" to each other correctly can take days of troubleshooting.

With Docker, we bypass both. The developer has already built the "perfect" computer for BUSCO; we just download it.

::: {.callout-important}

### **A Note on CPU Architectures**

If you are using an **Apple Silicon Mac (M1/M2/M3)** and try to pull an image built specifically for Intel/Windows chips, it might fail or run very slowly. You can force Docker to download the correct version or use emulation by adding the `--platform` flag:

```{bash dex}
#| eval: false
docker pull --platform linux/amd64 <image_name>
```

This flag can be used during both the **pull** and **run** steps.
:::

In the **Terminal**:

```{bash dpull}         
#| eval: false
# Pull the BUSCO image (a common genomics tool)
# Format: docker pull <author>/<tool>:<version>
docker pull ezlabgva/busco:v6.0.0_cv1
```

### **Running a Container**

Once the image is on your computer, you "run" it. This command is more complex because we have to tell Docker how to connect to our actual files.

```{bash drun}        
#| eval: false
# Breakdown of the command:
# --rm  : Automatically delete the container when the task is finished (saves space)
# -it   : Interactive Terminal (lets us see the output in real-time)
# -v    : Volume mapping. This connects your computer's folder to the container's folder.
#         Without this, the container cannot see your data!

docker run --rm -it ezlabgva/busco:v6.0.0_cv1 busco --help
```

::: {.callout-note}

### Micro-Challenge 4: Hello Docker

1.  **Check:** Open Docker Desktop. Look at the bottom left—it should say "Engine Running" (green).

2.  **Run:** Copy the `docker run` command above into your Terminal.

3.  **Verify:** Do you see the BUSCO help menu?

4.  **Document:** Now that you've run your first container, document this step in your `day3_reproducibility_notes.qmd` file. Add a secondary header, a brief explanation of what the command does, and a Bash code chunk (with `#| eval: false`). **Commit and Push** your notes to GitHub.

5.  **Logic:** If it worked, you just ran a high-end genomics tool inside a different operating system, without actually "installing" anything on your Mac or PC!
:::
