---
title: "Course 3, Day 3: Functional Genomics"
subtitle: "Genomic Variation & Variant Calling"
author: "Dr. Ido Hatam"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-copy: true
    self-contained: true
---

```{=html}
<style type="text/css">
body{ font-size: 18px; font-family: Arial, Helvetica, sans-serif; line-height: 30px; }
td { font-size: 14px; }
ul { font-size: 18px; }
h1.title { font-size: 38px; color: Black; }
h1 { font-size: 28px; color: Black; }
h2 { font-size: 22px; color: Black; }
h3 { font-size: 18px; font-family: Arial, Helvetica, sans-serif; color: Black; }
code.r{ font-size: 14px; }
.tocify { border: none; }
pre { font-size: 14px; }
</style>
```
# **Pre-Lab Checklist**

Before diving into the analysis, we must set up our workspace and ensure all dependencies are ready.

### **1. R Environment & Project Setup**

-   **Initialize Project:** Open RStudio and create a **New Project** in a new directory named `Course3_Day3_VariantCalling`. All subsequent steps assume you are working inside this project.

-   **Libraries:** Execute the setup chunk below to install/load the necessary packages.

```{r}
#| eval: false
#| label: setup

pkg_list <- c("VariantAnnotation", "GenomicFeatures", "GenomicRanges", "Gviz")

invisible(lapply(pkg_list, function(pkg) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    message(paste("Installing", pkg, "..."))
  
    if (!requireNamespace("BiocManager", quietly = TRUE)) {
      install.packages("BiocManager")
    }
    
    BiocManager::install(pkg, update = FALSE, ask = FALSE)
    library(pkg, character.only = TRUE)
    
  } else {
    message(paste(pkg, "is already installed and loaded."))
  }
}))
```

### **2. Data Management**

Now that your project is created, we will populate the `data/` folder with the necessary sequences.

-   **Reference Fasta:** Copy `S288C_reference.fasta` into your project's `data/` folder.

-   **Raw Reads:** Locate your original `reads.fastq` file from Course 2 and move/copy it to your project's `data/` folder.

-   **Reference GFF:** First, run the R script below to download the compressed GFF from NCBI.

```{r}
#| label: download_gff
#| eval: false

# Create data directory if it doesn't exist
dir.create("data", showWarnings = FALSE)

# NCBI URL for the S. cerevisiae S288C Reference GFF
gff_url <- "https://ftp.ncbi.nlm.nih.gov/genomes/refseq/fungi/Saccharomyces_cerevisiae/latest_assembly_versions/GCF_000146045.2_R64/GCF_000146045.2_R64_genomic.gff.gz"
dest_compressed <- "data/S288C_reference.gff.gz"

# Download the file using wget
message("Downloading reference GFF...")
download.file(url = gff_url, destfile = dest_compressed, method = "wget")
```

-   **Decompress the GFF:** Switch to the **Terminal** tab in RStudio and run:

```{bash}
#| eval: false
#| lebel: gzip
# Run this in your RStudio Terminal
gunzip -c data/S288C_reference.gff.gz > data/S288C_reference.gff
```

### **3. Software Check**

#### **A. Variant Calling Engine (Conda)**

-   **Docker:** Verify you have the medaka environment correctly set up 

```{bash}
#| eval: false
#| label: check
conda activate medaka_env
medaka -h
```

#### **B. Alignment & Processing Tools (Conda)**

In addition to `minimap2` and `samtools`, we need tools for handling VCFs and performing coordinate translations for tomorrow's isolate-centric analysis.

1.  **Check Environments:** In the Terminal, run `conda env list` to find your existing environments.

2.  **Create/Update Environment:** We will create a robust environment for variant analysis:

```{bash}
#| eval: false
#| label: create
# Create a dedicated environment with all necessary tools
# CrossMap is included for tomorrow's coordinate liftover
conda create -n variant_env -c bioconda -c conda-forge \
  minimap2 \
  samtools \
  bedtools \
  bcftools \
  crossmap \
  htslib

conda activate variant_env
```

# **Phase 1: Identifying Genetic Variation (Theory)**

### **1. Mapping vs. Assembly**

In Week 2, you performed *De Novo Assembly*. You reconstructed the genome by looking for overlaps between pieces—building the puzzle from scratch. This was necessary to see the physical "landscape" of your unique strain.

Today, we shift to **Reference-Based Mapping**. We treat the standard S288C genome as a blueprint and align our raw reads against it. This allows us to identify The Delta, or the specific points where your isolate diverges from the lab standard.

### **2. The Biology of Variation: Sequence vs. Structure**

Variation exists on a spectrum of scale. Identifying the right type of variation depends on how you look at your data.

#### **A. Single Nucleotide Polymorphisms (SNPs)**

SNPs are single base-pair changes (e.g., A $to$ T). They are the "fine print" of the genome. Because SNPs do not change the length or structure of the DNA, reference mapping is the perfect tool to identify them. We can easily see a single mismatch when a read is laid across the reference.

#### **B. Structural Variants (SVs): The "Keyhole" Problem**

Structural variants (SVs) involve 50 base pairs or more and include inversions, duplications, translocations, insertions, and deletions.

::: {.callout-important collapse="true"}
## **When mapping reads isn't enough**

If we *only* inspect raw reads by mapping them to a reference, we are looking at the genome through a keyhole. We're missing the genomic context surrounding the reads within the genome to which they belong.

-   Imagine an **inversion**: the sequence content is still there, so the reads will map to the reference. However, unless we have the *De Novo Assembly* (i.e. the context), we won't realize that a region along the chromosome has been flipped, potentially impacting gene expression.

-   Similarly, for **duplication**, mapping might show an increase in "Coverage" (more reads than expected), but it won't tell you *where* that second copy is located in the genome.

-   In addition, large **insertions** will not map to the reference and large **deletions** will be missing from the alignment. Without context, it will be impossible to tell where these insertions are located and if missing pieces are due to lack of coverage or a true deletion.

Mapping is excellent for finding "typos" (SNPs/smaller indels), especially when using long nanopore reads, but assembly and comparative genomics are required to truly understand "rearrangements" (SVs).
:::

### **3. The Broad Picture: Why Call Variants?**

-   **From Genetics to Genomics (A Holistic View):**

    -   **The Genetic Perspective:** Traditionally, researchers looked for specific mutations in single genes of interest (e.g., "Is *ADH1* broken?"). Calling SNPs and small indels allows us to perform this high-resolution "local" analysis, identifying if a single mutation in a gene like *ERG11* confers resistance to antifungal drugs.

    -   **The Genomic Perspective:** However, variant calling also allows us to assess the **entire biological system** at once—moving from the study of one gene to a **genome-scale analysis** of all pathways and their interactions. This provides a holistic view of:

        -   **Pathway Burden:** Instead of checking one enzyme, we can look at the mutational burden across *entire metabolic pathways* (e.g., Glycolysis).

        -   **System Tuning:** Variants in non-coding regions (promoters, UTRs) don't change the protein sequence, but they can drastically alter **gene expression levels**. This allows the organism to "tune" its metabolism genome-wide, assessing the functional load across the system rather than just individual parts.

-   **Strain Fingerprinting:** We use SNP profiles as high-resolution barcodes to distinguish between nearly identical strains (e.g., tracking contamination in a brewery).

-   **Pan-Genomics (Species-Wide Context):** A single Reference Genome (S288C) is just one individual. Pan-genomics allows us to see the "Accessory Genome"—genes that exist in wild strains but are missing in the lab standard. It also helps address **Reference Bias**, where reads from your unique strain fail to map because the reference simply doesn't contain that sequence.

    -   *Why aren't we building a Pan-Genome today?* While Pan-Genomics is the gold standard for structural context, it is extremely **computationally intensive** and **time-consuming**. It requires the *de novo* assembly and alignment of dozens of genomes simultaneously.

-   **Complex Traits & GWAS:** Most traits (like growth rate) are **Polygenic**. GWAS scans the *entire genome* to find statistical links between patterns of variants across multiple loci and specific metabolic capabilities.

-   **Phylogenetics:** Variants act as a molecular clock, allowing us to trace whether our wild strain is a "Sake" yeast or a "Wine" yeast.

**Our Focus Today:** Because we are using reference-based mapping, and due to the constraints of **compute power and time**, our primary focus will be on **SNPs and small Indels**. These are the variations where the S288C reference is still a reliable coordinate system and where we can generate results in a single lab session.

### **4. The Anatomy of the Evidence (File Formats)**

To do genomics, you must be able to read the raw evidence. We work with two primary "containers" for this information. Some of these you already worked with (e.g. BAM/SAM files), some are new.

#### **A. The Evidence Map: BAM (Binary Alignment Map)**

-   **Origin:** Generated by Aligners (e.g., `minimap2`).

-   **Purpose:** It records the placement of every single read on the reference genome.

-   **Inside the file:** Each line represents a read. It includes the read's sequence, the mapping quality (how sure are we that it belongs here?), and the **CIGAR string** (a code like `100M2D5M` that describes exactly where the read matches or has "gaps" relative to the reference).

-   **Usage:** We use the BAM to visualize the *support* for a variant. If a SNP is real, we should see it in many different reads at the same position.

#### **B. The Result: VCF (Variant Call Format) file**

-   **Origin:** Generated by **Variant Callers** (e.g., `Medaka`).

-   **Purpose:** This is the "Delta". It filters out the millions of bases that match the reference and only lists the sites that are different.

-   **The Anatomy of a VCF Record:**

| Column     | Description                                  | Example (SNP) | Example (Indel) |
|--------------|-------------------------------|--------------|--------------|
| **CHROM**  | The chromosome/contig name                   | NC_001133.9   | NC_001133.9     |
| **POS**    | The 1-based coordinate on the CHROM          | 31254         | 45902           |
| **REF**    | The base(s) in the **Reference** (S288C)     | A             | G               |
| **ALT**    | The base(s) in **Your Strain** (The Isolate) | T             | GAT             |
| **QUAL**   | Phred-scaled confidence score                | 60            | 45              |
| **INFO**   | Complex metadata (Depth, Tool stats)         | DP=45;AF=1.0  | DP=38;AF=0.5    |
| **FORMAT** | Definitions for the Sample column            | GT:DP:AD      | GT:DP:AD        |
| **SAMPLE** | The actual genotype of your yeast            | 1/1:45:0,45   | 0/1:38:18,20    |

**Understanding the Genotype (GT):**

-   `1/1`: **Homozygous Variant**. Both "copies" of the genome in your strain have the change.

-   `0/1`: **Heterozygous Variant**. One copy matches the reference (`0`), one copy has the variant (`1`). This is rare in haploid yeast but common in diploids!

# **Phase 2: The Demo (Pipeline Execution)**

To keep things light and ensure the analysis runs quickly during our lab session, we will process only **Chromosome 1** through a full variant calling pipeline. While the same logic applies to the entire genome, Chromosome 1 provides a manageable dataset to demonstrate the core principles of the "Delta."

### **Step 1: Alignment (The Map)**

-   **What:** We are mapping raw Nanopore reads to the Reference Chromosome 1.

-   **Why:** We cannot find variants in isolation. We must lay our reads across the "Standard" blueprint to identify exactly where our isolate diverges. `minimap2` performs the alignment, and `samtools` sorts the data into a binary format the computer can search quickly.

-   **Expected Output:** `aligned_chr1.bam` and its index `aligned_chr1.bam.bai`.

```{bash}
#| eval: false
#| label: align
# 1. Map reads to Reference
minimap2 -ax map-ont data/S288C_chr1.fasta data/reads.fastq | \
  samtools sort -o data/aligned_chr1.bam

# 2. Index the BAM (creates .bam.bai)
samtools index data/aligned_chr1.bam
```

### **Step 2: Variant Calling (The Judge)**

-   **What:** Running the `Medaka` variant calling engine.

-   **Why:** Raw alignments are messy. Some mismatches are real mutations, but others are just sequencing errors. `Medaka ` is a deep-learning tool that acts as a "judge"—it examines the pileup of reads and uses neural networks to determine if the evidence for a variant is strong enough to be called "Real."

-   **Expected Output:** A VCF file at `medaka/medaka.vcf`. Medaka also gives us a few additional files, but this is the only one we will be using today.

```{bash}
#| eval: false
#| label: medaka_variant

# Ensure medaka_env is active
medaka_variant -i data/reads.fastq -r data/S288C_chr1.fasta -t 8
```

### **Step 3: The Link (Did we break a gene?)**

-   **What:** Using `bedtools intersect` to compare our variant coordinates with gene coordinates.

    -   **Bedtools & BED Files:** `bedtools` is the industry-standard "Swiss Army Knife" for **Genomic Arithmetic**. While we are using it for VCFs and GFFs today, its namesake is the **BED (Browser Extensible Data)** format. A BED file is the simplest way to define a genomic region, requiring only three columns: `Chromosome`, `Start`, and `End`.

    -   **The Utility:** `bedtools` allows us to perform "spatial" questions across different formats. You can intersect a BAM with a BED to see read counts per region, or intersect a VCF with a GFF to find mutations in genes. It treats the genome like a set of overlapping shapes on a number line.

-   **Why:** A list of coordinates (e.g., `Chr1:4500`) is biologically meaningless on its own. By intersecting the VCF with the Reference GFF, we determine which variants physically overlap with annotated regions of the genome. To restrict the variants to genes, we will create a filtered gff that only has annotations on genes. This allows us to answer the critical question: *"Does this mutation actually impact a functional part of the genome?"*

-   **Expected Output:** A new VCF file `variants_in_genes.vcf` containing only the mutations that hit genes.


```{bash}
#| eval: false
#| label: intersect

#1 Create the gene only GFF
awk 'BEGIN{FS=OFS="\t"} $0 ~ /^#/ || $3=="gene"' \
data/S288C_reference.gff > data/S288C_genes.gff

#2 Check for overlaps between Variants (A) and Genes (B)
# Ensure variant_env is active
# We use -a (the variant file) and -b (the gene feature file)
bedtools intersect \
  -a medaka/medaka.vcf \
  -b data/S288C_genes.gff \
  -header > variants_in_genes.vcf
```

# **Phase 3: The "We Do" Visualization (The Gviz Stack)**

In this phase, we move into R to visualize our results. We want to see the evidence (BAM), the result (VCF), and the biological context (GFF) all in one place.

### **1. Data Ingestion**

-   **What:** We are loading the VCF file into R using `VariantAnnotation` and converting it to a `GRanges` object.

-   **Why:** The VCF is a text file. To plot it on a genomic axis, we need to transform it into a coordinate-based R object. We also filter for `PASS` variants to ensure we are only visualizing high-confidence calls.

-   **Expected Output:** An R object named `gr_variants` containing the high-quality genomic coordinates of your strain's mutations.

```{r}
#| eval: false
#| label: variant_ingest
vcf <- readVcf("variants_in_genes.vcf")
gr_variants <- rowRanges(vcf)
gr_variants <- gr_variants[fixed(vcf)$FILTER == "PASS"]
gr_variants <- gr_variants[seqnames(gr_variants) == "NC_001133.9"]
```

### **2. Track Construction**

-   **What:** Building the visual layers (tracks) for our plot.

-   **Why:** Genomic visualization is like a sandwich. We need layers for the reference genes (the blueprint), our identified variants (the delta), and the raw read coverage (the physical evidence). This allows us to see if a variant falls inside an exon or a promoter.

-   **Expected Output:** Three track objects: `gene_track`, `variant_track`, and `aln_track`.

```{r}
#| eval: false
#| label: tracks
# Build the Gene Blueprint from the GFF
txdb_ref <- makeTxDbFromGFF("data/S288C_reference.gff", format = "gff3")

gr_genes <- genes(txdb_ref)
gr_genes <- gr_genes[seqnames(gr_genes) == "NC_001133.9"]

gene_names <- mcols(gr_genes)$gene_id

names(gr_genes) <- gene_names
mcols(gr_genes)$gene <- names(gr_genes)   

gene_track <- GeneRegionTrack(
  gr_genes,
  chromosome = "NC_001133.9",
  name = "Gene Model",
  transcriptAnnotation = "gene",
  showID = TRUE
)

# Build the Delta Track (Your Variants)
variant_track <- AnnotationTrack(gr_variants,
                                chromosome = "NC_001133.9",
                                name = "Variants",
                                shape = "box",
                                fill  = "red"
)

# Build the Evidence Track (Your Reads)
aln_track <- AlignmentsTrack("data/aligned_chr1.bam",
                             chromosome = "NC_001133.9",
                             isPaired = FALSE, 
                             name = "Read Coverage")
```

### **3. The Reveal**

-   **What:** Drawing the final multi-track plot for a specific window of Chromosome 1.

-   **Why:** This is the moment of truth. We can visually verify the variant caller's decisions. Does the red box (variant) align with a clear gap or mismatch in the coverage track? Is it inside a gene?

-   **Expected Output:** A visualization in your RStudio Plots pane showing the relationship between reads, variants, and gene structures.

```{r}
#| eval: false
#| label: plot
plotTracks(list(GenomeAxisTrack(), gene_track, variant_track, aln_track),
           from = 30000, to = 32000, 
           chromosome = "NC_001133.9",
           main = "Variant Verification: Chr 1 Window")
```

# **Phase 4: Looking Ahead - Scaling to the Isolate Genome**

Tomorrow, we move from the "Sandbox" (Reference Chromosome 1) to an **Isolate-Centric** view.

### **1. The Full Pipeline Run**

We will apply today's logic to all 16 chromosomes simultaneously against the S288C reference. 

### **2. The Contextual Overlay (Mapping the Delta to Your Assembly)**

The goal of this course is to understand *your* organism. Tomorrow, we will "close the loop" by taking the variants you found today (relative to the S288C reference) and overlaying them against the **annotated assembly** you created in Week 2.

-   **Why perform an "Overlay"?**

    -   **Isolate-Centric View:** Instead of just asking "how is my yeast different from a lab standard?", we ask "how do these differences sit within the genomic landscape of *my* strain?"

    -   **Integrated Functional View:** By projecting reference-based SNPs onto your **structural annotations** from Day 1, you can see if the specific "Delta" you identified falls within the gene models you predicted on your own contigs.

    -   **Biological Context:** This allows you to visualize exactly where mutations sit relative to the unique architecture of your assembly—perhaps a SNP falls in a gene that is part of a structural rearrangement you discovered in Week 2.

**Get your `assembly_polished.fasta` and your annotated GFFs from Day 1 ready—tomorrow, we put the "Delta" into its true home.**
