---
title: "Lab 3: Genome Assembly and Annotation Basics"
---

<br>

# **The setup**

Create a new `GitHub` repo for this lab.

In RStudio:

- Create a new R Project with version control and connect it to your `GitHub` repo.
- Ensure your project contains a `data/` directory for raw inputs and an `outputs/` directory for generated results.
- Start a new Quarto document named `03-lab3.qmd`.
- Make sure to add the `data/` and `outputs/` folders to your `.gitignore` file.

Commit these files immediately with a message like `Initial commit`.

On your terminal, pull the Docker image for `quast`, which along with `BUSCO` will be used to evaluate the quality of the genome assembly you will do today. 

```{bash}
#| label: pull_quast
#| eval: false
docker pull quay.io/biocontainers/quast:5.3.0--py313pl5321h5ca1c30_2
```

Next, use `conda` to install the sequence aligner `minimap2` if not installed on your computer. 
You can check if it is installed by running `minimap2 -h` in your bash terminal. `minimap2` does not require a virtual environment. 

Also, load the following packages in your R session:

```{r}
#| label: packages
library("IRanges")
library("GenomicRanges")
library("Rsamtools")
library("GenomicAlignments")
library("Gviz")
```

Lastly, download the yeast reference genome, along with its GFF file (a file that contains the list of gene annotations) to be used for the assembly check, as well as the CDS containing FASTA file. This should be done in the terminal. 

**Remember to use `gunzip` to unzip the files that end in `.gz`.**  

```{bash}
#| label: refgen_dl
#| eval: false

# Download the reference genome from refseq
wget -nc -P data/ https://ftp.ncbi.nlm.nih.gov/genomes/refseq/fungi/Saccharomyces_cerevisiae/latest_assembly_versions/GCF_000146045.2_R64/GCF_000146045.2_R64_genomic.fna.gz

# Download the genome annotation file from refseq
wget -nc -P data/ https://ftp.ncbi.nlm.nih.gov/genomes/refseq/fungi/Saccharomyces_cerevisiae/latest_assembly_versions/GCF_000146045.2_R64/GCF_000146045.2_R64_genomic.gff.gz

# Download the cds containing fasta file from refseq
wget -nc -P data/ https://ftp.ncbi.nlm.nih.gov/genomes/refseq/fungi/Saccharomyces_cerevisiae/latest_assembly_versions/GCF_000146045.2_R64/GCF_000146045.2_R64_cds_from_genomic.fna.gz
```

::: {.callout-note}
**One final note before you begin:** the assignment questions for this lab are interspersed throughout the lab. There are 8 questions, make sure to answer all of them.
:::

***

# **Learning objectives**

<details>

<summary>Click to expand</summary>

**By the end of this lab you will be able to:**

- Understand the basic process and challenges of genome assembly.
- Understand the importance of polishing and scaffolding.
- Use Flye to perform long-read de novo genome assembly and Medaka to polish it.
- Evaluate assemblies using QUAST and BUSCO.
- Interpret common assembly quality metrics.
- Use GRanges for basic annotation-like tasks by mapping known genes to assembled contigs.

</details>

------------------------------------------------------------------------

# **Assembly overview**

A genome assembly is the process of piecing together the full DNA sequence from sequencing reads.
Long reads (like those from Oxford Nanopore) simplify this task because they provide more contiguous information. However, they 
also tend to have higher error rates compared to short reads.

In this lab, you'll assemble a small dataset of Oxford Nanopore reads from a _Saccharomyces cerevisiae_ isolate — the same strain you worked with in the previous lab. The dataset has low coverage, which will affect the quality of your assembly.

Important concepts:

- **Coverage** - How many times, on average, each position in the genome is represented by a read.
For example, 30× coverage means each base is represented, on average, by 30 different reads. High coverage improves the ability to generate a consensus sequence that corrects for sequencing or PCR errors. 

- **Contiguity (e.g., number of contigs, N50)** - A measure of how fragmented the assembly is.
One common metric is N50, which indicates that 50% of the assembled genome is contained in contigs that are at least that length. As a rule of thumb, a high N50 and a small number of contigs indicate a more contiguous (i.e., less fragmented) assembly.  

- **Completeness** - A measure of how much of the expected genome content is present in the assembly.
This is often assessed both by comparing the assembly size to expected size based on a reference genome, and whether key conserved genes (e.g., using `BUSCO`) are detected. For example, a complete assembly of _S. cerevisiae_ should contain nearly all expected genes for that species and be roughly 12Mb long. 

***

# **Assemble the genome**

You will assemble the genome using `Flye`, a _de novo_ assembler designed specifically for long noisy reads, such as those 
generated by Oxford Nanopore or PacBio sequencing platforms. It uses a repeat graph approach to assemble genomes, which allows it 
to handle complex repeat structures more effectively than traditional assemblers. `Flye` performs well on both low- and 
high-coverage datasets and is widely used in microbial, eukaryotic, and metagenomic projects.

::: {.callout-note}
A _de novo_ assembler is a tool that assembles sequencing reads into a genome without using a reference genome. It builds the 
sequence entirely from scratch, based only on overlaps and patterns in the reads themselves.
:::

```{bash}
#| label: flye
#| eval: false

# Run the Flye assembler, note the expected genome size
flye --nano-raw data/ERR1539006.fastq.gz --out-dir outputs/flye_output --threads 8 --genome-size 12m
```

* The `--nano-raw` flag denotes we are using raw nanopore reads with old sequencing chemistry.
* The `--threads` flag denotes number of CPU cores to use, here 8.
* The `--genome-size` flag is where the expected genome size is specified. Here 12Mb which is the expected size of the _S. cerevisiae_ genome. 

Once done, your assembled genome will be in `outputs/flye_output/assembly.fasta`, and a summary of the assembly process and stats will be in `outputs/flye_output/flye.log` but also outputted to terminal.

## **Assignment question 1**

In your document, have an R code chunk that tabulates the alignment statistics, make sure that the table shows the total length of the assembly, the N50, the number of contigs, largest contig and coverage. 
Based on that, would you expect to have a complete genome? Would you say that the coverage is adequate (3 marks)?

***

# **Cleaning up the mess: the concept of polishing**

*De novo* assembly, especially with noisy long reads, is a messy business. Even if the assembler successfully finds the correct *order* of the reads to build a contig, the individual letters (A, C, T, G) within that sequence might still be wrong due to sequencing errors. To fix this, we use polishing.

Think of polishing like running a spell-checker over a manuscript you just finished writing. The paragraphs are in the correct order, but there are typos everywhere. Polishing algorithms take the original reads, map them back to your newly built "draft" assembly, and look at the vertical stack of bases at every single position. If the majority of your reads disagree with the draft assembly at a certain spot, the polisher "outvotes" the draft and corrects the sequence.

Flye has its own built-in polishing algorithm that runs by default at the end of the assembly process. For expediency in this lab, we'll use just the default single round of Flye polishing. However, to truly get a high-quality consensus sequence, we need to bring in specialized tools. 

To take our assembly to the "Gold Standard," we will use Medaka. Medaka is a polisher developed directly by Oxford Nanopore. Unlike simple majority-voting algorithms, Medaka utilizes a Neural Network (AI) trained heavily on the physical properties of the nanopore itself. It recognizes and fixes the systematic error patterns inherent to the technology.

::: {.callout-warning}
**A note on our dataset:** Because we are working with a very old and noisy ONT dataset for this exercise, you will likely observe massive error rates reported during the Flye polishing round. I want to be entirely transparent with you: this is expected and is a direct consequence of the dataset's quality and age, not a flaw in your workflow.

For your own course project, working with modern chemistry, you should expect to run **5 rounds of Flye internal polishing** followed by **2 rounds of Medaka**. Additionally, for your own project data, you must carefully consult the Medaka documentation to select the appropriate model flag (`-m`), as this depends heavily on the specific ONT chemistry used for sequencing.
:::

## **Installing and running Medaka**

Let's set up Medaka using conda. 

::: {.callout-tip}
**Troubleshooting (The "Mac Lab" Fix)**
If you encounter errors creating the environment or running Medaka on our ARM64 lab computers, run the following to force osx-64 architecture for compatibility:

```bash
CONDA_SUBDIR=osx-64 conda create -n medaka_final -c conda-forge -c bioconda -c nanoporetech medaka python=3.10 -y
conda activate medaka_final
conda install -c conda-forge cffi -y
```
:::

Once installed, we will use it to polish our draft assembly. We will perform two rounds of polishing for this exercise.

```{bash}
#| label: medaka
#| eval: false

# Activate your environment
conda activate medaka_final

# Round 1
medaka_consensus -i data/ERR1539006.fastq.gz -d outputs/flye_output/assembly.fasta -o outputs/medaka_round1 -t 8

# Round 2
medaka_consensus -i data/ERR1539006.fastq.gz -d outputs/medaka_round1/consensus.fasta -o outputs/medaka_round2 -t 8
```

Your polished assembly will be located at `outputs/medaka_round2/consensus.fasta`. We will use this polished sequence going forward.

***

# **The final structure: scaffolding**

You've likely noticed that our assembly isn't just one continuous piece of DNA per chromosome. It's broken into multiple fragments called **contigs**. To turn these isolated islands of sequence into a finished, chromosome-level genome, we need a process called **scaffolding**.

Scaffolding is the process of figuring out the relative orientation and order of these contigs and connecting them together. Because we don't know the exact sequence of the DNA bridging the contigs, the assembler fills these gaps with strings of the letter **N**.

Due to the quality, age, and coverage of our example dataset, we will not be performing scaffolding in this lab exercise. The fragmentation is simply too high for it to be pedagogically useful here. 

However, do not ignore scaffolding! For your own course project, pending the quality and coverage of your initial assembly, I strongly redirect you to consider scaffolding tools (like RagTag) to elevate your assembly from a collection of contigs to a cohesive genome.

***

# **Evaluate the assembly**

You will evaluate the polished assembly using two tools: `quast` and `BUSCO`. `quast` is used to check and compare genome assemblies. It 
provides rudimentary metrics for contiguity such as N50 and the number of contigs, along with metrics that evaluate the accuracy of the assembly such as misassemblies.

`BUSCO` is used to evaluate the completeness of an assembly. It identifies the presence or absence of highly conserved, single-copy genes (orthologs) in the assembly, compared to a reference set for a specific lineage (in your case saccharomycetaceae). A high `BUSCO` score indicates a more complete representation of the expected gene content for a given organism.

The score is made of several components: 
- **C** - Level of completeness made up of: **S** (found in single copy as expected) and **D** (duplicated - appeared more than once. Could indicate assembly redundancy or real duplication.)
- **F** - Fragmented. Partial matches. May reflect incomplete assembly or gene breakage/partial deletions.
- **M** - Missing. Not found. Could be due to deletion, divergence, or poor assembly.

## **Evaluating the assembly using `BUSCO`**

Remember that `BUSCO` will be run using Docker, so make sure that Docker Desktop is active on your Mac.

```{bash}
#| label: busco
#| eval: false

docker run --rm \
  -v "$PWD":/data \
  -w /data \
  ezlabgva/busco:v6.0.0_cv1 \
  busco -i outputs/medaka_round2/consensus.fasta -o outputs/busco_output -l saccharomycetaceae_odb12 -m genome -c 8
```

* `-v "$PWD":/data` mounts the current working directory into the container as `/data`.
* `-w /data` ensures the container operates inside the mounted folder.
* `-c` represents the number of cores, in this case you will use 8.

::: {.callout-note}
`-l` is for lineage. Since we know the sample is a yeast (Saccharomyces), you will use the 
saccharomycetaceae_odb12 as reference. `BUSCO` also has the ability to try and identify the lineage of the 
submitted assembly using the `--auto-lineage-euk` flag. While useful, specifying the lineage is more reliable and saves runtime.
:::

### **Assignment question 2**

The `BUSCO` reports should be found under `outputs/busco_output`. Open the text file with lineage specific results, and answer the following:

* What were the `BUSCO` scores? Given the coverage do you think they are good (2 marks)?
* What was the N50 compared to the `flye` calculated one (1 mark)?
* What was the total assembly length (1 mark)?

## **Evaluating the assembly with `quast`**

`quast` will also be run using Docker. Before running QUAST, make sure you have unzipped your reference genomes.

```{bash}
#| label: unzip_refs
#| eval: false
gunzip data/GCF_000146045.2_R64_genomic.fna.gz
gunzip data/GCF_000146045.2_R64_genomic.gff.gz
```

```{bash}
#| label: quast
#| eval: false

docker run --rm -v "$(pwd)":/data -w /data quay.io/biocontainers/quast:5.3.0--py313pl5321h5ca1c30_2 \
  quast.py outputs/medaka_round2/consensus.fasta \
  -r data/GCF_000146045.2_R64_genomic.fna \
  -g data/GCF_000146045.2_R64_genomic.gff \
  -o outputs/quast_output
```

* The `-r` argument specifies the reference genome to which the assembly is compared. In this case, you will use the RefSeq reference genome for _S. cerevisiae_. A RefSeq genome is a curated, non-redundant, and annotated reference genome sequence maintained by NCBI.
* The `-g` argument specifies gene annotations by providing the GFF file for the reference genome. GFF stands for General Feature Format. It is a tab-delimited text file that describes genomic features.
* The `-o` argument specifies the output directory where `quast` will write its reports.

### **Assignment question 3**

Open the `quast` report:

* What was the genome fraction how does it compare to `BUSCO` (2 marks)?
* What was the duplication ratio (1 mark)?
* What was the level of misassemblies and what is the length of the misassembled contigs? What does it tell you about how well the reference genome represents the genomic diversity of the species _S. cerevisiae_ (3 marks)?

***

# **Annotating assemblies using alignment and visualization tools**

Next, you will align the contigs resulting from the genome assembly to the reference genome, and then align CDSs from the reference genome back to the contigs.

Aligning assembled contigs to a reference genome allows us to evaluate the structure and quality of the assembly. This step helps answer key questions like:
* Which parts of the genome were recovered?
* Are the contigs in the correct order and orientation?
* Do the contigs cover important genomic features, like genes or regulatory regions?

By mapping contigs to a known reference, we can also detect structural differences, identify potential misassemblies, and gain biological insights.

## **Why align the CDSs to the contig?**

Aligning the coding DNA sequences (CDSs) from a trusted reference genome to your assembled contigs allows you to annotate potential genes in your assembly. This helps identify what functional elements are present and where they are located. Even if your assembly is incomplete or fragmented, mapping known CDSs can reveal which parts of the genome were successfully recovered.

## **Assignment question 4: data prep**

For the next part you will work only with the sequence of chromosome 1, to reduce CPU time.

* Write a code chunk that extract the sequence of chromosome 1 from `data/GCF_000146045.2_R64_genomic.fna` and saves it as `outputs/R64_chr1.fna` (2 marks). 
* Make sure the chunk is on your Quarto file, and that you run it (1 mark). 
* Make sure the `R64_chr1.fna` is in your `.gitignore` as well (1 mark). 

## **Aligning your assembly to chromosome 1**

Next you will use `minimap2` to align the contigs to chromosome 1 extracted from the reference genome.
The `-ax asm5` argument is used when aligning closely related sequences, such as contigs to a reference genome, and requires a minimum of 95% sequence identity.  

```{bash}
#| label: chr1
#| eval: false

minimap2 -t 8 -ax asm5 outputs/R64_chr1.fna outputs/medaka_round2/consensus.fasta > outputs/contig_vs_chr1.sam
```

::: {.callout-note}
The output here is a `sam` (sequence alignment map) file, which is a tab delimited text file that specifies the coordinates of each aligned sequence relative to the reference.
:::

### **Using `samtools` to sort the alignment by region and converting it to a bam file**

In order to visualize the alignment, it needs to be sorted by coordinates so that fragments that are close to each other will indeed appear like that in the plot. The human-readable `sam` file will also be converted to a binary alignment map (`bam`) file. 

1. Convert the `sam` file to a `bam` file. 

```{bash}
#| label: samtobam
#| eval: false

samtools view -bS outputs/contig_vs_chr1.sam > outputs/alignment.bam
```

2. Sort the aligned contigs based on their coordinates along chromosome 1.

```{bash}
#| label: sortbat
#| eval: false

samtools sort outputs/alignment.bam -o outputs/alignment.sorted.bam
```

3. Index the sorted bam file. 

```{bash}
#| label: bam2bai
#| eval: false

samtools index outputs/alignment.sorted.bam
```

* Creates an index file (`.bai`) for the sorted BAM.
* Indexing allows fast access to specific regions, which is essential for genome browsers and R packages like `Gviz`.

### **Visualizing the alignment to chromosome 1 with GRanges**

Next you will visualize the aligned contigs against the chromosome using GRanges.

1. Import the `bam` file and convert it into a GRanges object.

```{r}
#| label: bamimport
#| eval: true

# BAM file path
bam_file <- "outputs/alignment.sorted.bam"

# Open connection to BAM file
bam <- BamFile(bam_file)

# Read all alignments
alns <- readGAlignments(bam)

# Convert to GRanges object
gr_alns <- granges(alns)
```

2. Create "tracks" showing the alignment.

```{r}
#| label: tracks
#| eval: true

aln_track <- AnnotationTrack(gr_alns, name = "Contigs", genome = "sacCer3", chromosome = "chr1")

# Add genome axis for scale
axis_track <- GenomeAxisTrack()
```

3. Visualize the alignment.

```{r}
#| label: viz1
#| eval: true

# Plot
plotTracks(list(axis_track, aln_track),
           from = min(start(gr_alns)),
           to = max(end(gr_alns)),
           main = "Contig Alignments to Chromosome 1")
```

***

> Congrats — you just built your own genome browser! While not as fancy or interactive as the UCSC genome browser, this R-based version is still incredibly powerful. It allows for customization, automation, and seamless integration with your analysis pipeline — all within a reproducible script.

***

#### **Assignment question 5**

- How many contigs did you have aligning to chromosome 1 (1 mark)?
- Make sure to have the plot in your document (1 mark).

### **Assignment question 6: garbing the longest contig**

Before proceeding with annotation:

* Write a code chunk that extracts the longest contig from your polished `consensus.fasta` file and saves it as `outputs/longest_contig.fasta`.
* The code is included in your assignment file.
* The resulting FASTA file is committed and pushed to your GitHub repository.

Why are we doing this?
Working with just the longest contig reduces computational load and simplifies the next steps, where you'll align known coding sequences (CDSs) to your assembly as a first pass at annotation.

### **Aligning CDSs to the longest contig**

1. Use `minimap2` to align the CDS containing fasta file to the longest contig. Remember to unzip it first.

```{bash}
#| label: unzip_cds
#| eval: false
gunzip data/GCF_000146045.2_R64_cds_from_genomic.fna.gz
```

```{bash}
#| label: longestvcds
#| eval: false

minimap2 -ax splice -uf --secondary=no outputs/longest_contig.fasta data/GCF_000146045.2_R64_cds_from_genomic.fna > outputs/cds_aln.sam
```

* The `-ax splice` argument uses the splice-aware preset, which is designed for aligning spliced transcripts (like mRNAs or CDSs) to a genome.
* The `-uf` flag indicates that the query sequences are full-length cDNA (or CDSs), which helps minimap2 model splicing more accurately.
* The `--secondary=no` argument suppresses secondary alignments.

### **Assignment question 7: getting bam and bai files for visualization**

* Write a code chunk that converts your `cds_aln.sam` file to a sorted and indexed `BAM` file for visualization (2 marks).
* Include the code in your document (1 mark).
* Ensure the final `BAM` is named `cds_aln.sorted.bam` (1 mark).
* Ensure that the Bam and Bai files are in the `.gitignore` file (1 mark). 
* Commit and push your changes to your GitHub repository (1 mark).

### **Assignment question 8: visualize CDS**

Using your sorted `BAM` file, create `Gviz` tracks that represent the aligned CDSs.

* Remember to read the alignment and convert it into GRanges.
* Remember working with annotation tracks.
* Remember to plot the alignment track.
* Remember to include your code and the plot in the document.
* Commit and push changes to `GitHub`.

***

# **You're all done**

Hope you had fun assembling and annotating genomes. This lab will become very useful once you start working on your project. Next lab will deal with transcriptomics data.